### 2017-10-02 ms.tex
adding files

\documentclass[useAMS, usenatbib, preprint, 12pt]{aastex}
\usepackage{cite, natbib}
\usepackage{float}
\usepackage{epsfig}
\usepackage{cases}
\usepackage[section]{placeins}
\usepackage{graphicx, subfigure}
\usepackage{color}
\usepackage{bm}

\newcommand{\columbia}{1}
\newcommand{\simons}{2}
\newcommand{\nyu}{3}
\newcommand{\cca}{4}
\newcommand{\mpia}{5}
\newcommand{\cds}{6}
\newcommand{\naigrain}{333}
\newcommand{\nmcquillan}{100}
\newcommand{\kepexample}{1430163}
\newcommand{\kepexampleperiod}{4}
\newcommand{\aigrainexampleperiod}{20.8}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Gaia}{{\it Gaia}}
\newcommand{\gaia}{{\it Gaia}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\acfRMS}{9.27}
\newcommand{\pgramRMS}{1.25}
\newcommand{\mcmcRMS}{0.46}
\newcommand{\racomment}[1]{{\color{red}#1}}

\begin{document}

\title{Investigating magnetic dynamo evolution with TESS field dwarfs}

\author{R. Angus\altaffilmark{\columbia, }\altaffilmark{\simons}
        J. R. Davenport, D. Busazi, D. Foreman-Mackey, A. Mann, S. Oh,
        D. Kipping
}

\altaffiltext{\columbia}{Department of Astronomy, Columbia University, NY, NY}
\altaffiltext{\simons}{Simons Fellow, RuthAngus@gmail.com}

\section{Introduction}

Among the key observable properties of main sequence stars in our Galaxy,
age is the most difficult to determine.
Traditionally, fitting isochrones to cluster stars was one of the only precise
methods for measuring ages but was extremely difficult for the majority of
isolated field stars, particularly for those without precise spectroscopic
information.
Methods such as asteroseismology and measuring Lithium abundances can provide
precise ages but require time intensive observations for each target and are
not capable of producing the large quantity of ages needed for exoplanet and
galactic population studies.

Galactic archaeology and exoplanet populations are two rapidly accelerating
fields of interest within astronomy. Although seemingly unconnected, these two
fields are linked by a mutual requirement for precise stellar parameters.
To galactic archaeologists, ages and elemental abundances are the most
important parameters.
Indeed, most galactic archaeology surveys target exactly these properties.
For exoplaneteers, masses and radii have historically been the most important
stellar parameters for understanding planetary systems.
With a growing number of planet hosts with precise masses and radii, attention
is turning toward other parameters such as ages to understand the history and
evolution of these systems.
Age is therefore a fundamental stellar parameter of great interest to two
large communities of astronomers.
However it is a difficult attribute to measure for main sequence F, G, K, and
M stars in the field, in part because low-mass dwarfs do not move far on the
Color-Magnitude diagram (CMD) during their hydrogen burning lifetimes.
Further, competing stellar evolution models predict different ages for the
same star.
Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age, and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics.

To improve our understanding of star and planet formation and evolution, as
well as the history of the Milky Way, we must constrain the ages of low-mass
stars like the Sun in the galactic field.
Fortunately nature has provided a powerful means to determine ages for main
sequence stars via their rotation.
Angular momentum is carried away though magnetically driven stellar winds,
which slows the star’s rotation over cosmic time. This rotation-based “clock”
is known as gyrochronology.
Cool spots on the star’s surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
While rotation periods have previously been measured from starspot-induced
flux modulations for hundreds of stars from the ground, space-based
photometric surveys have opened the door to homogeneous ensemble measures of
stellar rotation, and therefore age.
With precise light curves available from the TESS mission, we may be able to
determine rotation periods and ages for nearly 100,000 main sequence field
stars.

Additionally, rotation is related to stellar activity.
With so many TESS targets being M dwarfs which tend to be particularly active,
understanding the magnetic behavior of these stars has never been more
important.

To enable studies of stellar ages from rotation periods with TESS, we propose
to: 1. Measure accurate rotation periods for every available TESS FFI and
two-minute cadence target, using the most appropriate tools and methods
required.
2. Produce updated gyrochronology relations based on a wider range of field
star ages.
3. Determine the origin of the mysterious rotation period bimodality
discovered with Kepler by tracing the rotation period distribution over the
whole sky, and out to further distances utilizing public Gaia data.
4. Measure the star formation history across the sky using a new Bayesian
age-dating system.

\section{Scientific Justification}
\subsection{Rotation period bimodality}

\citet{McQuillan2014} demonstrated a striking discontinuity in the rotation
periods of Kepler M dwarfs.
They revealed a bimodal distribution of period, with a gap that depends on
mass.
Plenty of stars had rotation periods just shorter and just longer than this,
but very few had rotation periods around this value.
More recently, Davenport (2017) used Gaia data to elucidate this behaviour by
re-examining the Kepler rotation periods reported in McQuillan et al. (2014)
as a function of mass.
Davenport (2017) eliminated contaminating giants identified using Gaia
parallaxes, i.e. stars that are too bright to be
dwarfs given their distance.
These decontaminated data revealed that the period bimodality was not simply a
feature of M dwarfs but extended to the G and K stars.
Clearly then, this bimodal rotation period distribution is does not only
affect M dwarfs, but may be a feature of all stars, including the Sun.

\begin{figure}
\begin{center}
\includegraphics[width=3in, clip=true]{Davenport.png}
\caption{From Davenport (2017).
    Rotation period as a function of effective
temperature for the full McQuillan et al., (2014) Kepler sample in black, and
the subset of these stars that also feature in the TGAS Gaia DR1 catalogue in
blue. Contaminating giants have been removed from the blue sample and the
rotation period bimodality is revealed to exist across all temperatures shown.
    The red line is a 600 Myr rotational isochrone (also known as a gyrochrone).}
\label{fig:sip_hist}
\end{center}
\end{figure}

Intriguingly, the rotation gap is not constant across masses.
It appears at shorter rotation periods for more massive stars and longer
rotation periods for less massive stars.
In fact, it closely follows a contour of constant age, an isochrone of 600
Myr.
Two possible scenarios can explain this phenomenon.
The first is that stars undergo a transition in the behaviour of their
magnetic dynamo.
They switch from a fast to a slow rotating regime rapidly, spending little
time rotating at an intermittent period.
This transition would occur around 3-4 days for G dwarfs and 10-12 days for M
dwarfs.
Is this a convergence bump?
The rotation periods of stars are believed to have converged by around the age
of the Hyades, ~790 Myr (clusters younger than this show a broad spread in
rotation periods and older clusters a narrow one).
The age at which this transition occurs, around 600 Myr, is close to this.
Could this be a coincidence or do rotation periods undergo a rotation period
bump once they move stabilise onto a Skumanich spin-down regime?
If this is the case, a study of this transition could greatly advance our
understanding of the operation of stellar dynamos.

The rotation period bimodality seen in McQuillan et al. (2014) and Davenport
(2017) could also be explained by two waves of star formation: a short one
centered around a few hundred million years ago and a long one centered around
a few billion years ago.
TESS’s full view of the sky will allow us to establish the spatial extent of
this rotation period gap and investigate the underlying cause.

\subsection{Gyrochronology}

Gyrochronology is an active area of interest in astronomy today due to its applicability to stellar, exoplanet and galactic astronomy.
It is a method of age-dating stars that doesn’t require expensive
spectroscopy, requiring only a rotation period and a colour or temperature.
Considering the vast number of light curves that will be made available by
future photometric surveys, as well as the already formidable number produced
by the Kepler spacecraft, gyrochronology is by far the most widely applicable,
relatively precise age-dating method available.
Although the phenomenon of magnetic braking has been closely studied since its
discovery in the mid/late 20th century, the manifestations of magnetic fields
and their effects on rotation and activity are complex; the physics
underpinning rotational evolution are far from well-understood.
In fact, concerning rotation, though Kepler was a triumphant planet hunter and
a revolutionary stellar mission, it may have revealed as many new mysteries as
it resolved old ones.
One mystery is described above; it is the appearance of a bimodal period
distribution.
The other concerns gyrochronology, the method of inferring the age of a star
from its rotation period and mass.

Although the classical spin-down law of Skumanich (1972), Period $\propto$
Age$^1/2$ holds for all open clusters with measured periods, it does not
appear to describe old field stars (Angus et al., 2015, van Saders et al.,
2016).
Asteroseismic pulsators observed by Kepler, older than the Sun, rotate more
rapidly than the Skumanich law predicts they should. Van Saders et al. (2016)
identified a mass-dependence in the… recalibrated the gyrochronology
relations, accounting for this divergence from the fiducial spin-down
relation.
They attribute the difference between old and new models to weakened magnetic
braking.
Once a star reaches a critical Rossby number (the ratio of the rotation period
to the convective overturn time), its magnetic dynamo transitions into a
different state which drastically reduces (basically switches off) braking
efficiency.
This means a 5 Gyr star of Solar mass will have the same rotation period as a
6 Gyr and rotation period cannot be used as an age proxy.
The transition occurs at around Solar age for Solar mass stars but later for
less massive stars and earlier for more massive stars.
So, although this effect reduces the usefulness of gyrochronology for stars
more massive than the Sun, in practise, for Solar masses and below,
gyrochronology can still be considered an effective dating method.

We intend to use the rotation periods of planet hosting stars to infer the
age-dependence of planet frequency. Kepler alone does not produce sufficient
stars with rotation periods to reliably characterise the time-dependent
frequency of exoplanets.
Mann et al. (2016a), Mann et al. (2016b) and Rizzuto et al. (2017) detected
transiting exoplanets in four young open clusters observed by the K2 mission.

\section{Measuring Rotation Periods}
In this era of large photometric surveys (Kepler/K2, TESS, WFIRST, LSST, PTF,
PanStarrs and more) rotation periods are quickly becoming one of the most
accessible properties of stars.
Precise light curves produced by these surveys often reveal the presence of
dark spots on the surfaces of cool stars which revolve with the stellar
surface creating an overall dimming effect once every rotation period.
Dark spotted regions and bright faculae leave a characteristic trace in the
light curve from which a rotation period can be inferred.
The extraction of a rotation period from a light curve can be as
straight-forward as computing a Lomb-Scargle periodogram or, in cases where
the signal is less clear, can be inferred via modelling the correlation
between data points.
This latter approach could involve either computing an autocorrelation
function (McQuillan et al., 2013), or fitting a Gaussian process to the time
series (Angus et al., 2017, Foreman-Mackey et al., 2017).
Signals produced by the rotation of spotted stars can have amplitudes of a few
percent of the total flux, but can also have very low amplitudes of a few
parts per million.
The frequencies of high amplitude signals are easy to measure and, in these
cases, most measured frequencies will agree regardless of the technique used
to measure them.
Similarly, short-period signals are easy to measure, especially if light curve
variations are sinusoidal in shape.
In the low amplitude and long period cases however, care is needed to separate
real astrophysical signal from instrumental systematics.

\begin{figure}
\begin{center}
\includegraphics[width=6in, clip=true]{Kalesalad.png}
\caption{All stars observed during K2 campaign 4, plotted according to their
equatorial coordinates and colored by their preliminary rotation period.
These rotation periods were measured using a simple ACF method, applied to
    everest (Luger et al., 2015) light curves.}
\label{fig:kalesalad}
\end{center}
\end{figure}

We intend to measure rotation periods for all mid to low-mass dwarfs that show
evidence of rotational modulation in their light curves.

Need to measure rotation periods of all stars if you want to infer the
distribution of planets as a function of age.

It looks like hotter stars stop spinning down at around the age of the Sun.
However, that does not necessarily mean that gyrochronology cannot be used to
infer ages.
Gyro works for low mass stars down to around 0.35 M$_\odot$ (CITATION), up to
the age of the Universe.
It works for Solar-mass stars up to the age of the Sun and for stars more
massive than the Sun (but less massive than the Kraft break) it works to
gradually decreasing ages.
This means that X\% of stars in the TIC are likely to have ages that can be
determined by gyrochronology.

\section{Analysis Plan}

\section{Technical Feasibility}

\section{Expected Impact}

\section{Budget Justification}

\bibliographystyle{plainnat}
\bibliography{references}
\end{document}

### 2017-10-02 ms.tex
adding bits from the ADAP

@@
 
-\citet{McQuillan2014} demonstrated a striking discontinuity in the rotation
-periods of Kepler M dwarfs.
-They revealed a bimodal distribution of period, with a gap that depends on
-mass.
-Plenty of stars had rotation periods just shorter and just longer than this,
-but very few had rotation periods around this value.
-More recently, Davenport (2017) used Gaia data to elucidate this behaviour by
-re-examining the Kepler rotation periods reported in McQuillan et al. (2014)
-as a function of mass.
-Davenport (2017) eliminated contaminating giants identified using Gaia
-parallaxes, i.e. stars that are too bright to be
-dwarfs given their distance.
-These decontaminated data revealed that the period bimodality was not simply a
-feature of M dwarfs but extended to the G and K stars.
-Clearly then, this bimodal rotation period distribution is does not only
-affect M dwarfs, but may be a feature of all stars, including the Sun.
+One of the most remarkable results from the Kepler rotation period catalog was
+the discovery of a bimodal period distribution among field stars by McQuillan
+et al. (2013), which is shown
+in Figure \ref{davenport}.
+While a separate sequence of rapid rotation periods had been known in young
+stellar clusters due to lower-mass stars settling on to the angular momentum
+main sequence slower \citep[\eg][]{barnes2007}, this new bimodality was
+detected from slower rotating M dwarfs (T_$\mathrm{eff} <$ 4000).
+The bimodality separated M dwarfs into two populations, those with rotation
+periods longer than $\sim$ 20 days, and those with periods between $\sim$ 1
+and 20 days.
+Follow-up analysis of the Kepler data by \citet{mcquillan2014} found the
+period bimodality extended to include K dwarfs.
+Recently, Co-I Davenport discovered this period bimodality extends throughout
+all masses in the Kepler rotation sample for nearby stars, as shown in Figure
+\ref{davenport} \citep{davenport2017}.
+Two formation scenarios have been proposed to explain the observed period
+bimodality.
+The first scenario, initially proposed by \citet{mcquillan2013}, is the
+rotation period distribution reflects the local star formation history, and
+thus the bimodality represents a drop in the star formation rate around 600
+Myr ago.
+This model is supported by both the extension of the bimodality to earlier
+spectral types by \citet{davenport2017}, and also the tentative detection that
+the two rotation period populations have distinct proper motion distributions.
+However, such a variation in the star formation rate on short timescales has
+some tension with independent observational efforts to determine the local
+star formation history.
+While Color–Magnitude diagram inversions from Hipparcos have suggested a
+similarly short timescale variation in star formation of $\sim$ 0.5 Gyr
+\citep{hernandez2000}, other studies find slower variations over several
+Gyr \citep[\eg][]{cignoni2006}.
+Using white dwarf cooling models to infer the local formation history
+(“cosmochronology”) also supports higher star formation several Gyr ago, but
+can rarely achieve age resolution better than $\sim$ 1 Gyr due to small sample
+sizes \citep{tremblay2014}.
+The spatial extent of such coherent and localized variations in star formation
+history is unknown.
+The second scenario to explain this feature is that the period bimodality
+occurs due to a previously unknown variation in the spin-down evolution for
+low-mass stars.
+In this model the star formation history would be continuous over the past
+$\sim$ 1 Gyr, and around 600 Myr stars would move quickly through the observed
+period minima due to this unknown phase transition or feedback mechanism.
+While this model is not predicted by angular momentum loss prescriptions,
+rapid transitions in rotation period are observed for stars in young clusters.
+Stars move quickly from the rapidly rotating “convective” sequence (periods of
+$\sim<1$ day) to the “interface” sequence (periods of several days) during the
+first few hundred Myr, with lower mass stars taking longer to make this
+transition as they settle onto the main sequence \citep{barnes2003}.
+Secondly, a gap in chromospheric activity levels for solar-type field stars
+has also been observed \citep{vaughan1980}.
+While this magnetic activity indicator smoothly varies with stellar ages over
+long timescales, the gap indicates a rapid transition phase from “active” to
+“inactive” is present within the first Gyr \citep{pace2009}.
+Thirdly, the angular momentum loss underpinning gyrochronology seems to slow
+for stars older than the Sun, indicating a potential change in the magnetic
+dynamo for slower rotators \citep{angus2015, vansaders2016}.
+The K2 rotation period sample provides the ideal dataset to test these two
+formation scenarios.
+If the bimodality is due to an age distribution we would expect to only see
+the feature locally, and that it could disappear at further distances or along
+different lines of sight where small scale variations in the star formation
+history are less apparent.
+Coherent changes in the rotation period (or stellar age) distribution along
+lines of sight might also reveal details of the spiral arm pattern speed near
+the Sun.
+The kinematic separation between the two rotation period populations would be
+reinforced by supplemental measurements from the upcoming public Gaia data
+releases.
+However, if the bimodality is truly due to a transition point in the spin-down
+evolution at young ages, we would expect to find no little to no variation in
+this feature as seen in Figure 2b with galactic position, distance, or between
+K2 fields.
+lines of sight might also reveal details of the spiral arm pattern speed near
+the Sun.
+The kinematic separation between the two rotation period populations would be
+reinforced by supplemental measurements from the upcoming public Gaia data
+releases.
+However, if the bimodality is truly due to a transition point in the spin-down
+evolution at young ages, we would expect to find no little to no variation in
+this feature as seen in Figure 2b with galactic position, distance, or between
+K2 fields.
+
+% \citet{McQuillan2014} demonstrated a striking discontinuity in the rotation
+% periods of Kepler M dwarfs.
+% They revealed a bimodal distribution of period, with a gap that depends on
+% mass.
+% Plenty of stars had rotation periods just shorter and just longer than this,
+% but very few had rotation periods around this value.
+% More recently, Davenport (2017) used Gaia data to elucidate this behaviour by
+% re-examining the Kepler rotation periods reported in McQuillan et al. (2014)
+% as a function of mass.
+% Davenport (2017) eliminated contaminating giants identified using Gaia
+% parallaxes, i.e. stars that are too bright to be
+% dwarfs given their distance.
+% These decontaminated data revealed that the period bimodality was not simply a
+% feature of M dwarfs but extended to the G and K stars.
+% Clearly then, this bimodal rotation period distribution is does not only
+% affect M dwarfs, but may be a feature of all stars, including the Sun.
 
@@
     The red line is a 600 Myr rotational isochrone (also known as a gyrochrone).}
-\label{fig:sip_hist}
+\label{fig:davenport}
 \end{center}

### 2017-10-02 ms.tex
a few words about photometry

--- 2017-10-02 ms.tex
@@
 
+The measurement of rotation periods is less sensitive to crowding and source
+confusion than exoplanet transit characterization because the rotation period
+is not effected by photometric dilution.
+We will apply two complementary methods to extract and calibrate light curves
+from the TESS FFIs.
+First, for bright or isolated targets, we will follow \citet{montet2017} to
+estimate aperture shapes and perform aperture photometry for bright sources.
+Using an ensemble of sources, we will de-trend these light curves using a
+modified version of \textsf{everest} \citep{luger2016, luger2017} designed to
+preserve photometric signatures of rotation.
+This will be achieved by fitting for the systematic effects in the light curve
+using the \textsf{everest} model simultaneously with a Gaussian Process model
+for the astrophysical variation.
+Both \citet{aigrain2016} and \citet{luger2016} demonstrated that this can
+preserve stellar variability signals and we will use the \textsf{celerite}
+algorithm \citep{dfm2017} to scale the computations to the size of TESS FFI
+datasets.
+
+The precision of existing photometric de-trending methods degrades in crowded
+fields \citep[for example,][]{luger2017}.
+However, to make robust measurements of rotation periods, we do not need
+absolutely calibrated light curves.
+Therefore, in crowded fields, we will apply a difference imaging method that
+was developed for the K2 Campaign 9 microlensing project \citep{henderson2016}
+based on the \textsf{CPM} \citep{wang2016} to robustly measure the photometric
+variations of crowded sources.
+Unlike standard difference imaging methods, this procedure does not require a
+reference image.
+Instead, a causal data-driven model is built to predict the time series in
+every pixel taking systematic effects into account and the residuals between
+the observations and the model predictions provide an estimate of the
+astrophysical variability in each pixel.
+We will tune this method preserve rotation signals and apply it to detect
+rotation across the FFIs.
+
 \begin{figure}

### 2017-10-02 ms.tex

--- 2017-10-02 ms.tex
@@
 main sequence slower \citep[\eg][]{barnes2007}, this new bimodality was
-detected from slower rotating M dwarfs (T_$\mathrm{eff} <$ 4000).
+detected from slower rotating M dwarfs (T$_\mathrm{eff} <$ 4000).
 The bimodality separated M dwarfs into two populations, those with rotation
@@
 
+The measurement of rotation periods is less sensitive to crowding and source
+confusion than exoplanet transit characterization because the rotation period
+is not effected by photometric dilution.
+We will apply two complementary methods to extract and calibrate light curves
+from the TESS FFIs.
+First, for bright or isolated targets, we will follow \citet{montet2017} to
+estimate aperture shapes and perform aperture photometry for bright sources.
+Using an ensemble of sources, we will de-trend these light curves using a
+modified version of \textsf{everest} \citep{luger2016, luger2017} designed to
+preserve photometric signatures of rotation.
+This will be achieved by fitting for the systematic effects in the light curve
+using the \textsf{everest} model simultaneously with a Gaussian Process model
+for the astrophysical variation.
+Both \citet{aigrain2016b} and \citet{luger2016} demonstrated that this can
+preserve stellar variability signals and we will use the \textsf{celerite}
+algorithm \citep{dfm2017} to scale the computations to the size of TESS FFI
+datasets.
+
+The precision of existing photometric de-trending methods degrades in crowded
+fields \citep[for example,][]{luger2017}.
+However, to make robust measurements of rotation periods, we do not need
+absolutely calibrated light curves.
+Therefore, in crowded fields, we will apply a difference imaging method that
+was developed for the K2 Campaign 9 microlensing project \citep{henderson2016}
+based on the \textsf{CPM} \citep{wang2016} to robustly measure the photometric
+variations of crowded sources.
+Unlike standard difference imaging methods, this procedure does not require a
+reference image.
+Instead, a causal data-driven model is built to predict the time series in
+every pixel taking systematic effects into account and the residuals between
+the observations and the model predictions provide an estimate of the
+astrophysical variability in each pixel.
+We will tune this method preserve rotation signals and apply it to detect
+rotation across the FFIs.
+
 \begin{figure}

### 2017-10-03 ms.tex
some changes to the text

--- 2017-10-02 ms.tex
@@
 
-Among the key observable properties of main sequence stars in our Galaxy,
-age is the most difficult to determine.
-Traditionally, fitting isochrones to cluster stars was one of the only precise
-methods for measuring ages but was extremely difficult for the majority of
-isolated field stars, particularly for those without precise spectroscopic
-information.
-Methods such as asteroseismology and measuring Lithium abundances can provide
-precise ages but require time intensive observations for each target and are
-not capable of producing the large quantity of ages needed for exoplanet and
-galactic population studies.
+% Among the key observable properties of main sequence stars in our Galaxy,
+% age is the most difficult to determine.
+% Traditionally, fitting isochrones to cluster stars was one of the only precise
+% methods for measuring ages but was extremely difficult for the majority of
+% isolated field stars, particularly for those without precise spectroscopic
+% information.
+% Methods such as asteroseismology and measuring Lithium abundances can provide
+% precise ages but require time intensive observations for each target and are
+% not capable of producing the large quantity of ages needed for exoplanet and
+% galactic population studies.
 
 Galactic archaeology and exoplanet populations are two rapidly accelerating
-fields of interest within astronomy. Although seemingly unconnected, these two
-fields are linked by a mutual requirement for precise stellar parameters.
+fields of interest within astronomy.
+Although seemingly unconnected, these fields are linked by a mutual
+requirement for precise stellar parameters.
 To galactic archaeologists, ages and elemental abundances are the most
@@
 evolution of these systems.
-Age is therefore a fundamental stellar parameter of great interest to two
-large communities of astronomers.
+Age is therefore a fundamental stellar parameter of great interest to at least
+two large communities of astronomers.
 However it is a difficult attribute to measure for main sequence F, G, K, and
@@
 leverage for advancing our knowledge of galactic archeology as well as
-exoplanet population demographics.
+exoplanet population demographics via gyrochronology.
 
-To improve our understanding of star and planet formation and evolution, as
-well as the history of the Milky Way, we must constrain the ages of low-mass
-stars like the Sun in the galactic field.
-Fortunately nature has provided a powerful means to determine ages for main
-sequence stars via their rotation.
 Angular momentum is carried away though magnetically driven stellar winds,
-which slows the star’s rotation over cosmic time. This rotation-based “clock”
+which slows the star’s rotation over cosmic time. This rotation-based `clock'
 is known as gyrochronology.
@@
 stellar rotation, and therefore age.
-With precise light curves available from the TESS mission, we may be able to
-determine rotation periods and ages for nearly 100,000 main sequence field
-stars.
+Using precise light curves available from the TESS mission, we hope to
+infer ages for nearly 100,000 main sequence field stars, including thousands
+of planet hosts.
+The ages of the planet hosts will allow us to characterize trends in planet
+properties over time, by adding to the \kepler\ sample.
 
@@
 age-dating system.
+5. Further characterize the intrinsic uncertainty in gyrochronology ages using
+comoving pairs identified in the first Gaia data release.
 
@@
 One of the most remarkable results from the Kepler rotation period catalog was
-the discovery of a bimodal period distribution among field stars by McQuillan
-et al. (2013), which is shown
-in Figure \ref{davenport}.
+the discovery of a bimodal period distribution among field stars by
+\citet{mcquillan2013}, which is shown in Figure \ref{davenport}.
 While a separate sequence of rapid rotation periods had been known in young
@@
 dynamo for slower rotators \citep{angus2015, vansaders2016}.
-The K2 rotation period sample provides the ideal dataset to test these two
+The TESS rotation period sample provides the ideal dataset to test these two
 formation scenarios.
@@
 evolution at young ages, we would expect to find no little to no variation in
-this feature as seen in Figure 2b with galactic position, distance, or between
-K2 fields.
-lines of sight might also reveal details of the spiral arm pattern speed near
-the Sun.
-The kinematic separation between the two rotation period populations would be
-reinforced by supplemental measurements from the upcoming public Gaia data
-releases.
-However, if the bimodality is truly due to a transition point in the spin-down
-evolution at young ages, we would expect to find no little to no variation in
-this feature as seen in Figure 2b with galactic position, distance, or between
-K2 fields.
+this feature as seen in Figure \ref{fig:davenport} with galactic position,
+distance, or across the sky.
 
@@
 
-Intriguingly, the rotation gap is not constant across masses.
-It appears at shorter rotation periods for more massive stars and longer
-rotation periods for less massive stars.
-In fact, it closely follows a contour of constant age, an isochrone of 600
-Myr.
-Two possible scenarios can explain this phenomenon.
-The first is that stars undergo a transition in the behaviour of their
-magnetic dynamo.
-They switch from a fast to a slow rotating regime rapidly, spending little
-time rotating at an intermittent period.
-This transition would occur around 3-4 days for G dwarfs and 10-12 days for M
-dwarfs.
-Is this a convergence bump?
-The rotation periods of stars are believed to have converged by around the age
-of the Hyades, ~790 Myr (clusters younger than this show a broad spread in
-rotation periods and older clusters a narrow one).
-The age at which this transition occurs, around 600 Myr, is close to this.
-Could this be a coincidence or do rotation periods undergo a rotation period
-bump once they move stabilise onto a Skumanich spin-down regime?
-If this is the case, a study of this transition could greatly advance our
-understanding of the operation of stellar dynamos.
+% Intriguingly, the rotation gap is not constant across masses.
+% It appears at shorter rotation periods for more massive stars and longer
+% rotation periods for less massive stars.
+% In fact, it closely follows a contour of constant age, an isochrone of 600
+% Myr.
+% Two possible scenarios can explain this phenomenon.
+% The first is that stars undergo a transition in the behaviour of their
+% magnetic dynamo.
+% They switch from a fast to a slow rotating regime rapidly, spending little
+% time rotating at an intermittent period.
+% This transition would occur around 3-4 days for G dwarfs and 10-12 days for M
+% dwarfs.
+% Is this a convergence bump?
+% The rotation periods of stars are believed to have converged by around the age
+% of the Hyades, ~790 Myr (clusters younger than this show a broad spread in
+% rotation periods and older clusters a narrow one).
+% The age at which this transition occurs, around 600 Myr, is close to this.
+% Could this be a coincidence or do rotation periods undergo a rotation period
+% bump once they move stabilise onto a Skumanich spin-down regime?
+% If this is the case, a study of this transition could greatly advance our
+% understanding of the operation of stellar dynamos.
 
-The rotation period bimodality seen in McQuillan et al. (2014) and Davenport
-(2017) could also be explained by two waves of star formation: a short one
-centered around a few hundred million years ago and a long one centered around
-a few billion years ago.
-TESS’s full view of the sky will allow us to establish the spatial extent of
-this rotation period gap and investigate the underlying cause.
+% The rotation period bimodality seen in McQuillan et al. (2014) and Davenport
+% (2017) could also be explained by two waves of star formation: a short one
+% centered around a few hundred million years ago and a long one centered around
+% a few billion years ago.
+% TESS’s full view of the sky will allow us to establish the spatial extent of
+% this rotation period gap and investigate the underlying cause.
 
@@
 
-Gyrochronology is an active area of interest in astronomy today due to its applicability to stellar, exoplanet and galactic astronomy.
-It is a method of age-dating stars that doesn’t require expensive
-spectroscopy, requiring only a rotation period and a colour or temperature.
-Considering the vast number of light curves that will be made available by
-future photometric surveys, as well as the already formidable number produced
-by the Kepler spacecraft, gyrochronology is by far the most widely applicable,
-relatively precise age-dating method available.
+Gyrochronology is by far the most widely applicable, relatively precise
+age-dating method available.
 Although the phenomenon of magnetic braking has been closely studied since its
@@
 underpinning rotational evolution are far from well-understood.
-In fact, concerning rotation, though Kepler was a triumphant planet hunter and
-a revolutionary stellar mission, it may have revealed as many new mysteries as
-it resolved old ones.
-One mystery is described above; it is the appearance of a bimodal period
-distribution.
-The other concerns gyrochronology, the method of inferring the age of a star
-from its rotation period and mass.
-
-Although the classical spin-down law of Skumanich (1972), Period $\propto$
-Age$^1/2$ holds for all open clusters with measured periods, it does not
-appear to describe old field stars (Angus et al., 2015, van Saders et al.,
-2016).
+Although the classical spin-down law of \citet{skumanich1972}: Period
+$\propto$ Age$^1/2$ holds for all open clusters with measured periods, it does
+not appear to describe old field stars \citep{angus2015, vansaders2016}.
 Asteroseismic pulsators observed by Kepler, older than the Sun, rotate more
-rapidly than the Skumanich law predicts they should. Van Saders et al. (2016)
-identified a mass-dependence in the… recalibrated the gyrochronology
-relations, accounting for this divergence from the fiducial spin-down
-relation.
-They attribute the difference between old and new models to weakened magnetic
-braking.
-Once a star reaches a critical Rossby number (the ratio of the rotation period
-to the convective overturn time), its magnetic dynamo transitions into a
-different state which drastically reduces (basically switches off) braking
+rapidly than the \citet{skumanich1972} law predicts they should.
+\citet{vansaders2016} find that including a transition to a weakened magnetic
+braking regime in the gyrochronology models, at a critical Rossby number,
+provides an improved fit to the data.
+Once a star reaches a critical Rossby number (the ratio of the rotation
+period to the convective overturn time), its magnetic dynamo transitions into
+a different state which drastically reduces (basically switches off) braking
 efficiency.
@@
 gyrochronology can still be considered an effective dating method.
+
+Since Solar-type magnetic dynamos are generated through a combination of
+convective motion and rotation, gyrochronology is thought to only be
+applicable to stars with significant convective envelopes (T$_{\mathrm{eff}}
+\lesssim$ 6250 K) that are not yet fully convective (3250 $\lesssim \mathrm{T}
+_{\mathrm{eff}}$).
 

### 2017-10-03 ms.tex
trying to change bib

@@
 
-\newcommand{\columbia}{1}
-\newcommand{\simons}{2}
-\newcommand{\nyu}{3}
-\newcommand{\cca}{4}
-\newcommand{\mpia}{5}
-\newcommand{\cds}{6}
-\newcommand{\naigrain}{333}
-\newcommand{\nmcquillan}{100}
-\newcommand{\kepexample}{1430163}
-\newcommand{\kepexampleperiod}{4}
-\newcommand{\aigrainexampleperiod}{20.8}
 \newcommand{\Kepler}{{\it Kepler}}
@@
 \newcommand{\numax}{$\nu_{\mathrm{max}}$}
-\newcommand{\acfRMS}{9.27}
-\newcommand{\pgramRMS}{1.25}
-\newcommand{\mcmcRMS}{0.46}
+
 \newcommand{\racomment}[1]{{\color{red}#1}}
+
+\newcommand{\columbia}{1}
+\newcommand{\simons}{2}
+\newcommand{\ww}{3}
+\newcommand{\nsf}{4}
+\newcommand{\florida}{5}
+\newcommand{\cca}{6}
+\newcommand{\hubble}{7}
+\newcommand{\princeton}{8}
 
@@
 
-\author{R. Angus\altaffilmark{\columbia, }\altaffilmark{\simons}
-        J. R. Davenport, D. Busazi, D. Foreman-Mackey, A. Mann, S. Oh,
-        D. Kipping
+\author{R. Angus\altaffilmark{\columbia}
+% \altaffilmark{\simons}
+        J. R. Davenport\altaffilmark{\ww}
+% \altaffilmark{\nsf},
+        D. Busazi\altaffilmark{\florida},
+        D. Foreman-Mackey\altaffilmark{\cca},
+        A. Mann\altaffilmark{\columbia}
+% \altaffilmark{\hubble},
+        S. Oh\altaffilmark{\princeton},
+        D. Kipping\altaffilmark{\columbia}
 }
 
-\altaffiltext{\columbia}{Department of Astronomy, Columbia University, NY, NY}
+\altaffiltext{\columbia}{Department of Astronomy, Columbia University, New
+York, NY}
 \altaffiltext{\simons}{Simons Fellow, RuthAngus@gmail.com}
+\altaffiltext{\ww}{Western Washington University, Bellingham, WA}
+\altaffiltext{\nsf}{NSF Fellow}
+\altaffiltext{\florida}{Florida Gulf Coast University, Fort Myers, FL}
+\altaffiltext{\cca}{Center for Computational Astronomy, Flatiron Institute,
+New York, NY}
+\altaffiltext{\hubble}{Hubble Fellow}
+\altaffiltext{\princeton}{Princeton University, Princeton, NJ}
 
 \section{Introduction}
-
-% Among the key observable properties of main sequence stars in our Galaxy,
-% age is the most difficult to determine.
-% Traditionally, fitting isochrones to cluster stars was one of the only precise
-% methods for measuring ages but was extremely difficult for the majority of
-% isolated field stars, particularly for those without precise spectroscopic
-% information.
-% Methods such as asteroseismology and measuring Lithium abundances can provide
-% precise ages but require time intensive observations for each target and are
-% not capable of producing the large quantity of ages needed for exoplanet and
-% galactic population studies.
 
@@
 With a growing number of planet hosts with precise masses and radii, attention
-is turning toward other parameters such as ages to understand the history and
-evolution of these systems.
+is turning toward other parameters, such as ages, to understand the history
+and evolution of these systems.
 Age is therefore a fundamental stellar parameter of great interest to at least
 two large communities of astronomers.
-However it is a difficult attribute to measure for main sequence F, G, K, and
-M stars in the field, in part because low-mass dwarfs do not move far on the
-Color-Magnitude diagram (CMD) during their hydrogen burning lifetimes.
+Unfortunately however, it is a difficult attribute to measure for main sequence
+F, G, K, and M stars in the field, in part because low-mass dwarfs do not move
+far on the Color-Magnitude diagram (CMD) during their hydrogen burning
+lifetimes.
 Further, competing stellar evolution models predict different ages for the
@@
 Angular momentum is carried away though magnetically driven stellar winds,
-which slows the star’s rotation over cosmic time. This rotation-based `clock'
-is known as gyrochronology.
+which slows the star’s rotation over cosmic time.
+This rotation-based `clock' is known as gyrochronology.
 Cool spots on the star’s surface rotate in-to and out-of view, creating small
@@
 The ages of the planet hosts will allow us to characterize trends in planet
-properties over time, by adding to the \kepler\ sample.
+properties over time.
 
-Additionally, rotation is related to stellar activity.
+Additionally, as rotation both influences and is influenced {\it by} stellar
+magnetic fields, it is inextricably related to stellar activity.
 With so many TESS targets being M dwarfs which tend to be particularly active,
@@
 
-To enable studies of stellar ages from rotation periods with TESS, we propose
-to: 1. Measure accurate rotation periods for every available TESS FFI and
-two-minute cadence target, using the most appropriate tools and methods
-required.
+To enable studies of stellar ages and magnetic activity from rotation periods
+with TESS, we propose to:
+1. Measure accurate rotation periods for every available TESS FFI and
+two-minute cadence target.
 2. Produce updated gyrochronology relations based on a wider range of field
@@
 for the astrophysical variation.
-Both \citet{aigrain2016b} and \citet{luger2016} demonstrated that this can
+Both \citet{aigrain2016} and \citet{luger2016} demonstrated that this can
 preserve stellar variability signals and we will use the \textsf{celerite}
@@
 
-\section{Analysis Plan}
-
-\section{Technical Feasibility}
-
 \section{Expected Impact}
+We will provide light curves and rotation periods for both the two-minute
+cadence and FFI targets.
 
 \section{Budget Justification}
+PI Angus intends to use the budget to employ a student for 3-4 months.
+In this time the student will assist in extracting light curves from the FFIs,
+measuring rotation periods and building a rotation period catalog.
+The student will also be involved in the scientific project of their choosing:
+either the rotation period bimodality, gyrochronology of low-mass stars, or
+gyrochronology of comoving pairs.
+
+% \begin{thebibliography}
+% \bibitem[Aigrain et al.(2016)]{aigrain2016} Aigrain, S., Parviainen,
+%     H., \& Pope, B.~J.~S.\ 2016, \mnras, 459, 2408
+% \bibitem[Angus et al.(2015)]{angus2015} Angus, R., Aigrain, S.,
+% Foreman-Mackey, D., \& McQuillan, A.\ 2015, \mnras, 450, 1787
+% \bibitem[Barnes(2003)]{barnes2003} Barnes, S.~A.\ 2003, \apj, 586,
+% \bibitem[Barnes(2007)]{barnes2007} Barnes, S.~A.\ 2007, \apj, 669, 1167
+% \bibitem[Cignoni et al.(2006)]{cignoni2006} Cignoni, M., Degl'Innocenti, S.,
+% Prada Moroni, P.~G., \& Shore, S.~N.\ 2006, \aap, 459, 783 464
+% \bibitem[Davenport(2017)]{davenport2017} Davenport, J.~R.~A.\ 2017, \apj, 835,
+% 16
+% \bibitem[Henderson et al.(2016)]{henderson2016} Henderson, C.~B.,
+% Poleski, R., Penny, M., et al.\ 2016, \pasp, 128, 124401
+% \bibitem[Hernandez et al.(2000)]{hernandez2000} Hernandez, X.,
+% Valls-Gabaud, D., \& Gilmore, G.\ 2000, \mnras, 316, 605
+% \bibitem[Luger et al.(2016)]{luger2016} Luger, R., Agol, E., Kruse, E., et
+%     al.\ 2016, \aj, 152, 100
+% \bibitem[Luger et al.(2017)]{luger2017} Luger, R., Kruse, E., Foreman-Mackey,
+%     D., Agol, E., \& Saunders, N.\ 2017, arXiv:1702.05488
+% \bibitem[Foreman-Mackey et al.(2017)]{foreman-mackey2017} Foreman-Mackey, D.,
+%     Agol, E., Ambikasaran, S., \& Angus, R.\ 2017, arXiv:1703.09710
+% \bibitem[McQuillan et al.(2013)]{mcquillan2013} McQuillan, A., Aigrain, S., \&
+%     Mazeh, T.\ 2013, \mnras, 432, 1203
+% \bibitem[McQuillan et al.(2014)]{Mcquillan2014} McQuillan, A., Mazeh, T., \&
+%     Aigrain, S.\ 2014, \apjs, 211, 24
+% \bibitem[Montet et al.(2017)]{montet2017} Montet, B.~T., Tovar, G., \&
+%     Foreman-Mackey, D.\ 2017, arXiv:1705.07928
+% \bibitem[Pace et al.(2009)]{pace2009} Pace, G., Melendez, J., Pasquini, L., et
+%     al.\ 2009, \aap, 499, L9
+% \bibitem[Skumanich(1972)]{skumanich1972} Skumanich, A.\ 1972, \apj, 171, 565
+% \bibitem[Tremblay et al.(2014)]{tremblay2014} Tremblay, P.-E., Kalirai, J.~S.,
+%     Soderblom, D.~R., Cignoni, M., \& Cummings, J.\ 2014, \apj, 791, 92
+% \bibitem[Vaughan \& Preston(1980)]{vaughan1980} Vaughan, A.~H., \& Preston,
+%     G.~W.\ 1980, \pasp, 92, 385
+% \bibitem[Wang et al.(2016)]{wang2016} Wang, D., Hogg, D.~W., Foreman-Mackey,
+%     D., \& Sch{\"o}lkopf, B.\ 2016, \pasp, 128, 094503
+% \end{thebibliography}
 
@@
 \bibliography{references}
+
 \end{document}

### 2017-10-03 ms.tex
ref styles
aggghhhhhh not enough time

@@
 \usepackage{bm}
+
+\bibliographystyle{aasjournal}
 
@@
 the discovery of a bimodal period distribution among field stars by
-\citet{mcquillan2013}, which is shown in Figure \ref{davenport}.
+\citet{mcquillan2013}, which is shown in Figure \ref{fig:davenport}.
 While a separate sequence of rapid rotation periods had been known in young
@@
 all masses in the Kepler rotation sample for nearby stars, as shown in Figure
-\ref{davenport} \citep{davenport2017}.
+\ref{fig:davenport} \citep{davenport2017}.
 Two formation scenarios have been proposed to explain the observed period
@@
 for stars older than the Sun, indicating a potential change in the magnetic
-dynamo for slower rotators \citep{angus2015, vansaders2016}.
-The TESS rotation period sample provides the ideal dataset to test these two
+dynamo for slower rotators \citep{angus2015, van-saders2016}.
+
+The TESS FFI and CTL targets will provide the ideal dataset to test these two
 formation scenarios.
-If the bimodality is due to an age distribution we would expect to only see
-the feature locally, and that it could disappear at further distances or along
+If the bimodality is due to an age distribution we would expect only to see
+the feature locally; it could disappear at greater distances or along
 different lines of sight where small scale variations in the star formation
 history are less apparent.
-Coherent changes in the rotation period (or stellar age) distribution along
-lines of sight might also reveal details of the spiral arm pattern speed near
-the Sun.
-The kinematic separation between the two rotation period populations would be
-reinforced by supplemental measurements from the upcoming public Gaia data
-releases.
 However, if the bimodality is truly due to a transition point in the spin-down
 evolution at young ages, we would expect to find no little to no variation in
-this feature as seen in Figure \ref{fig:davenport} with galactic position,
-distance, or across the sky.
+this feature as seen in Figure \ref{fig:davenport} with galactic position.
 
-% \citet{McQuillan2014} demonstrated a striking discontinuity in the rotation
-% periods of Kepler M dwarfs.
-% They revealed a bimodal distribution of period, with a gap that depends on
-% mass.
-% Plenty of stars had rotation periods just shorter and just longer than this,
-% but very few had rotation periods around this value.
-% More recently, Davenport (2017) used Gaia data to elucidate this behaviour by
-% re-examining the Kepler rotation periods reported in McQuillan et al. (2014)
-% as a function of mass.
-% Davenport (2017) eliminated contaminating giants identified using Gaia
-% parallaxes, i.e. stars that are too bright to be
-% dwarfs given their distance.
-% These decontaminated data revealed that the period bimodality was not simply a
-% feature of M dwarfs but extended to the G and K stars.
-% Clearly then, this bimodal rotation period distribution is does not only
-% affect M dwarfs, but may be a feature of all stars, including the Sun.
+To map the rotation period distribution as a function of galactocentric
+position we will match the TESS FFI and CTL targets to the upcoming data
+release from the Gaia mission \citep{perryman2001}.
+This will also allow us to remove contaminating subgiants and binary stars
+from the sample, leaving only single main sequence stars.
+The period distribution of G dwarfs is critical for this analysis since they
+are observable at greater distances, allowing us to sample different star
+formation histories.
+With the April 2018 data release from Gaia (DR2) estimate that we will be able
+to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$ 3 kpc.
+Unlike \Kepler, since \TESS\ is targeting the whole sky, it will be possible
+to investigate the rotation period distribution as a function of
+galactocentric position.
+The size scale over which the Milky Way’s star formation history varies is
+unknown however significant variations in star formation histories are seen in
+Andromeda between 100 pc volumes, as well as large galaxy-wide trends
+\citep[\eg][]{lewis2015}.
 
@@
 
-% Intriguingly, the rotation gap is not constant across masses.
-% It appears at shorter rotation periods for more massive stars and longer
-% rotation periods for less massive stars.
-% In fact, it closely follows a contour of constant age, an isochrone of 600
-% Myr.
-% Two possible scenarios can explain this phenomenon.
-% The first is that stars undergo a transition in the behaviour of their
-% magnetic dynamo.
-% They switch from a fast to a slow rotating regime rapidly, spending little
-% time rotating at an intermittent period.
-% This transition would occur around 3-4 days for G dwarfs and 10-12 days for M
-% dwarfs.
-% Is this a convergence bump?
-% The rotation periods of stars are believed to have converged by around the age
-% of the Hyades, ~790 Myr (clusters younger than this show a broad spread in
-% rotation periods and older clusters a narrow one).
-% The age at which this transition occurs, around 600 Myr, is close to this.
-% Could this be a coincidence or do rotation periods undergo a rotation period
-% bump once they move stabilise onto a Skumanich spin-down regime?
-% If this is the case, a study of this transition could greatly advance our
-% understanding of the operation of stellar dynamos.
-
-% The rotation period bimodality seen in McQuillan et al. (2014) and Davenport
-% (2017) could also be explained by two waves of star formation: a short one
-% centered around a few hundred million years ago and a long one centered around
-% a few billion years ago.
-% TESS’s full view of the sky will allow us to establish the spatial extent of
-% this rotation period gap and investigate the underlying cause.
-
 \subsection{Gyrochronology}
@@
 $\propto$ Age$^1/2$ holds for all open clusters with measured periods, it does
-not appear to describe old field stars \citep{angus2015, vansaders2016}.
+not appear to describe old field stars \citep{angus2015, van-saders2016}.
 Asteroseismic pulsators observed by Kepler, older than the Sun, rotate more
 rapidly than the \citet{skumanich1972} law predicts they should.
-\citet{vansaders2016} find that including a transition to a weakened magnetic
+\citet{van-saders2016} find that including a transition to a weakened magnetic
 braking regime in the gyrochronology models, at a critical Rossby number,
 provides an improved fit to the data.
-Once a star reaches a critical Rossby number (the ratio of the rotation
-period to the convective overturn time), its magnetic dynamo transitions into
-a different state which drastically reduces (basically switches off) braking
-efficiency.
+Once a star reaches a critical Rossby number (the ratio of the rotation period
+to the convective overturn time), magnetic braking efficiency effectively
+drops to zero.
 This means a 5 Gyr star of Solar mass will have the same rotation period as a
@@
 less massive stars and earlier for more massive stars.
-So, although this effect reduces the usefulness of gyrochronology for stars
-more massive than the Sun, in practise, for Solar masses and below,
-gyrochronology can still be considered an effective dating method.
-
-Since Solar-type magnetic dynamos are generated through a combination of
-convective motion and rotation, gyrochronology is thought to only be
-applicable to stars with significant convective envelopes (T$_{\mathrm{eff}}
-\lesssim$ 6250 K) that are not yet fully convective (3250 $\lesssim \mathrm{T}
-_{\mathrm{eff}}$).
+Although this effect reduces the usefulness of gyrochronology for stars more
+massive than the Sun, in practise, for Solar-mass stars, down to M dwarfs with
+masses greater than 0.35$M_\odot$, gyrochronology can still be considered an
+effective dating method.
+This will include a large number of \TESS\ planet candidates.
 
 We intend to use the rotation periods of planet hosting stars to infer the
-age-dependence of planet frequency. Kepler alone does not produce sufficient
-stars with rotation periods to reliably characterise the time-dependent
-frequency of exoplanets.
+age-dependence of planet frequency.
+Kepler alone has not yet produced a sufficient number of stars with rotation
+periods to reliably characterise the time-dependent frequency of exoplanets.
 Mann et al. (2016a), Mann et al. (2016b) and Rizzuto et al. (2017) detected
@@
 
-% \begin{thebibliography}
-% \bibitem[Aigrain et al.(2016)]{aigrain2016} Aigrain, S., Parviainen,
-%     H., \& Pope, B.~J.~S.\ 2016, \mnras, 459, 2408
-% \bibitem[Angus et al.(2015)]{angus2015} Angus, R., Aigrain, S.,
-% Foreman-Mackey, D., \& McQuillan, A.\ 2015, \mnras, 450, 1787
-% \bibitem[Barnes(2003)]{barnes2003} Barnes, S.~A.\ 2003, \apj, 586,
-% \bibitem[Barnes(2007)]{barnes2007} Barnes, S.~A.\ 2007, \apj, 669, 1167
-% \bibitem[Cignoni et al.(2006)]{cignoni2006} Cignoni, M., Degl'Innocenti, S.,
-% Prada Moroni, P.~G., \& Shore, S.~N.\ 2006, \aap, 459, 783 464
-% \bibitem[Davenport(2017)]{davenport2017} Davenport, J.~R.~A.\ 2017, \apj, 835,
-% 16
-% \bibitem[Henderson et al.(2016)]{henderson2016} Henderson, C.~B.,
-% Poleski, R., Penny, M., et al.\ 2016, \pasp, 128, 124401
-% \bibitem[Hernandez et al.(2000)]{hernandez2000} Hernandez, X.,
-% Valls-Gabaud, D., \& Gilmore, G.\ 2000, \mnras, 316, 605
-% \bibitem[Luger et al.(2016)]{luger2016} Luger, R., Agol, E., Kruse, E., et
-%     al.\ 2016, \aj, 152, 100
-% \bibitem[Luger et al.(2017)]{luger2017} Luger, R., Kruse, E., Foreman-Mackey,
-%     D., Agol, E., \& Saunders, N.\ 2017, arXiv:1702.05488
-% \bibitem[Foreman-Mackey et al.(2017)]{foreman-mackey2017} Foreman-Mackey, D.,
-%     Agol, E., Ambikasaran, S., \& Angus, R.\ 2017, arXiv:1703.09710
-% \bibitem[McQuillan et al.(2013)]{mcquillan2013} McQuillan, A., Aigrain, S., \&
-%     Mazeh, T.\ 2013, \mnras, 432, 1203
-% \bibitem[McQuillan et al.(2014)]{Mcquillan2014} McQuillan, A., Mazeh, T., \&
-%     Aigrain, S.\ 2014, \apjs, 211, 24
-% \bibitem[Montet et al.(2017)]{montet2017} Montet, B.~T., Tovar, G., \&
-%     Foreman-Mackey, D.\ 2017, arXiv:1705.07928
-% \bibitem[Pace et al.(2009)]{pace2009} Pace, G., Melendez, J., Pasquini, L., et
-%     al.\ 2009, \aap, 499, L9
-% \bibitem[Skumanich(1972)]{skumanich1972} Skumanich, A.\ 1972, \apj, 171, 565
-% \bibitem[Tremblay et al.(2014)]{tremblay2014} Tremblay, P.-E., Kalirai, J.~S.,
-%     Soderblom, D.~R., Cignoni, M., \& Cummings, J.\ 2014, \apj, 791, 92
-% \bibitem[Vaughan \& Preston(1980)]{vaughan1980} Vaughan, A.~H., \& Preston,
-%     G.~W.\ 1980, \pasp, 92, 385
-% \bibitem[Wang et al.(2016)]{wang2016} Wang, D., Hogg, D.~W., Foreman-Mackey,
-%     D., \& Sch{\"o}lkopf, B.\ 2016, \pasp, 128, 094503
-% \end{thebibliography}
-
-\bibliographystyle{plainnat}
 \bibliography{references}

### 2017-10-03 ms.tex
still ridiculously long, this isn't funny

@@
 Using precise light curves available from the TESS mission, we hope to
-infer ages for nearly 100,000 main sequence field stars, including thousands
-of planet hosts.
+infer ages for nearly 100,000 main sequence field stars, including a few
+hundred planet hosts.
 The ages of the planet hosts will allow us to characterize trends in planet
@@
 two-minute cadence target.
-2. Produce updated gyrochronology relations based on a wider range of field
-star ages.
-3. Determine the origin of the mysterious rotation period bimodality
+2. Determine the origin of the mysterious rotation period bimodality
 discovered with Kepler by tracing the rotation period distribution over the
 whole sky, and out to further distances utilizing public Gaia data.
-4. Measure the star formation history across the sky using a new Bayesian
+3. Measure the star formation history across the sky using a new Bayesian
 age-dating system.
-5. Further characterize the intrinsic uncertainty in gyrochronology ages using
+4. Further characterize the intrinsic uncertainty in gyrochronology ages using
 comoving pairs identified in the first Gaia data release.
@@
 effective dating method.
-This will include a large number of \TESS\ planet candidates.
 
-We intend to use the rotation periods of planet hosting stars to infer the
-age-dependence of planet frequency.
-Kepler alone has not yet produced a sufficient number of stars with rotation
-periods to reliably characterise the time-dependent frequency of exoplanets.
+{\bf We will use the rotation periods of TESS field dwarfs to calibrate the
+gyrochronology relations using galactic kinematics}.
+Many stars in the TESS CTL will also have proper motions, parallaxes,
+positions and radial velocities published in the second \Gaia\ data release.
+These parameters provide the information necessary to calculate both
+galactocentric positions and action angles of the stars, both of which are age
+indicators.
+Vertical action (angular momentum through the plane of the Milky Way)
+increases over time due to orbital heating and has been demonstrated to
+increase with age \citep[\eg][]{west2011}.
+Galactocentric position can also determine a star's age, since different
+stellar populations, thin or thick disk, halo and bulge have different average
+ages.
+PI Angus is currently using the \kepler\ rotation periods of stars in the
+first \Gaia\ data release to calibrate the relations between rotation period
+and vertical action dispersion, two tracers of age.
+Since this relation is likely vary with both galactic longitude and latitude,
+the rotation periods produced from \TESS\ light curves will enable a
+comprehensive, all-sky calibration of these relations.
+Since one of the few ways to accurately age-date fully convective, late M
+dwarfs is via kinematics, these new relations will help to infer ages for all
+stars with M$<$ 0.35M$_\odot$.
+
+{\bf We will use the rotation periods of planet hosting stars to infer
+the age-dependence of planet frequency.}j Kepler alone has not yet produced a
+sufficient number of stars with rotation periods to reliably characterise the
+time-dependent frequency of exoplanets.
 Mann et al. (2016a), Mann et al. (2016b) and Rizzuto et al. (2017) detected

### 2017-10-03 ms.tex
adding pdf

@@
 \author{R. Angus\altaffilmark{\columbia}
-% \altaffilmark{\simons}
         J. R. Davenport\altaffilmark{\ww}
-% \altaffilmark{\nsf},
+        D. Foreman-Mackey\altaffilmark{\cca},
+        S. Oh\altaffilmark{\princeton},
         D. Busazi\altaffilmark{\florida},
-        D. Foreman-Mackey\altaffilmark{\cca},
         A. Mann\altaffilmark{\columbia}
-% \altaffilmark{\hubble},
-        S. Oh\altaffilmark{\princeton},
         D. Kipping\altaffilmark{\columbia}
@@
 
-Galactic archaeology and exoplanet populations are two rapidly accelerating
-fields of interest within astronomy.
-Although seemingly unconnected, these fields are linked by a mutual
-requirement for precise stellar parameters.
-To galactic archaeologists, ages and elemental abundances are the most
-important parameters.
-Indeed, most galactic archaeology surveys target exactly these properties.
-For exoplaneteers, masses and radii have historically been the most important
-stellar parameters for understanding planetary systems.
-With a growing number of planet hosts with precise masses and radii, attention
-is turning toward other parameters, such as ages, to understand the history
-and evolution of these systems.
-Age is therefore a fundamental stellar parameter of great interest to at least
-two large communities of astronomers.
-Unfortunately however, it is a difficult attribute to measure for main sequence
-F, G, K, and M stars in the field, in part because low-mass dwarfs do not move
-far on the Color-Magnitude diagram (CMD) during their hydrogen burning
-lifetimes.
-Further, competing stellar evolution models predict different ages for the
-same star.
+Age is a fundamental stellar parameter of great interest to at galactic
+archaeologists and exoplaneteers alike but unfortunately it is a difficult
+attribute to measure for main sequence F, G, K, and M stars in the field.
 Of the measurable properties for a large ensemble of field stars, rotation
@@
 stellar rotation, and therefore age.
-Using precise light curves available from the TESS mission, we hope to
-infer ages for nearly 100,000 main sequence field stars, including a few
-hundred planet hosts.
-The ages of the planet hosts will allow us to characterize trends in planet
-properties over time.
+Using precise light curves available from the TESS mission, we expect to
+measure the rotation periods of around ... stars.
+Many of these rotation periods will be convertible into ages as they will be
+intermediate age FGK and early M dwarfs.
+Of particular interest will be the expected ~300 gyrochronal ages of exoplanet
+host stars.
+The gyrochronal ages of these planet hosts, combined with their isochronal and
+kinematic ages will allow us to characterize trends in planet properties over
+time and across the galaxy.
 
@@
 magnetic fields, it is inextricably related to stellar activity.
-With so many TESS targets being M dwarfs which tend to be particularly active,
-understanding the magnetic behavior of these stars has never been more
+With so many \TESS\ targets being M dwarfs which tend to be particularly
+active, understanding the magnetic behavior of these stars has never been more
 important.
 
-To enable studies of stellar ages and magnetic activity from rotation periods
-with TESS, we propose to:
+We propose to:
 1. Measure accurate rotation periods for every available TESS FFI and
@@
 comoving pairs identified in the first Gaia data release.
+5. Infer the distribution of exoplanets as a function of time and galactic
+position.
 
@@
 \citet{mcquillan2013}, which is shown in Figure \ref{fig:davenport}.
-While a separate sequence of rapid rotation periods had been known in young
-stellar clusters due to lower-mass stars settling on to the angular momentum
-main sequence slower \citep[\eg][]{barnes2007}, this new bimodality was
-detected from slower rotating M dwarfs (T$_\mathrm{eff} <$ 4000).
-The bimodality separated M dwarfs into two populations, those with rotation
+The bimodality separats M dwarfs into two populations, those with rotation
 periods longer than $\sim$ 20 days, and those with periods between $\sim$ 1
@@
 \ref{fig:davenport} \citep{davenport2017}.
-Two formation scenarios have been proposed to explain the observed period
-bimodality.
-The first scenario, initially proposed by \citet{mcquillan2013}, is the
+Two scenarios have been proposed to explain the observed period bimodality.
+In the first scenario, initially proposed by \citet{mcquillan2013}, the
 rotation period distribution reflects the local star formation history, and
@@
 This model is supported by both the extension of the bimodality to earlier
-spectral types by \citet{davenport2017}, and also the tentative detection that
-the two rotation period populations have distinct proper motion distributions.
-However, such a variation in the star formation rate on short timescales has
-some tension with independent observational efforts to determine the local
-star formation history.
-While Color–Magnitude diagram inversions from Hipparcos have suggested a
+spectral types by \citet{davenport2017}, and also the tentative evidence for
+two distinct proper motion distributions.
+However, such a variation in the star formation rate on short timescales is in
+tension with independent observational efforts to determine the local star
+formation history.
+While color–magnitude diagram inversions from Hipparcos have suggested a
 similarly short timescale variation in star formation of $\sim$ 0.5 Gyr
@@
 history is unknown.
-The second scenario to explain this feature is that the period bimodality
-occurs due to a previously unknown variation in the spin-down evolution for
-low-mass stars.
-In this model the star formation history would be continuous over the past
+The second scenario explaining this feature is a previously unknown variation
+in the spin-down evolution for low-mass stars.
+In this model, the star formation history would be continuous over the past
 $\sim$ 1 Gyr, and around 600 Myr stars would move quickly through the observed
-period minima due to this unknown phase transition or feedback mechanism.
+period minima due to some phase transition or feedback mechanism.
 While this model is not predicted by angular momentum loss prescriptions,
@@
 The TESS FFI and CTL targets will provide the ideal dataset to test these two
-formation scenarios.
-If the bimodality is due to an age distribution we would expect only to see
-the feature locally; it could disappear at greater distances or along
-different lines of sight where small scale variations in the star formation
-history are less apparent.
+scenarios explaining the appearance of a period bimodality.
+If the bimodal period distribution reflects a age discontinuous age
+distribution, the feature should be local; it could
+disappear at greater distances or along different lines of sight.
 However, if the bimodality is truly due to a transition point in the spin-down
-evolution at young ages, we would expect to find no little to no variation in
-this feature as seen in Figure \ref{fig:davenport} with galactic position.
+evolution at young ages, there should be little to no variation in the feature
+with galactic position.
 
@@
 from the sample, leaving only single main sequence stars.
-The period distribution of G dwarfs is critical for this analysis since they
-are observable at greater distances, allowing us to sample different star
-formation histories.
-With the April 2018 data release from Gaia (DR2) estimate that we will be able
-to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$ 3 kpc.
-Unlike \Kepler, since \TESS\ is targeting the whole sky, it will be possible
-to investigate the rotation period distribution as a function of
-galactocentric position.
-The size scale over which the Milky Way’s star formation history varies is
-unknown however significant variations in star formation histories are seen in
-Andromeda between 100 pc volumes, as well as large galaxy-wide trends
-\citep[\eg][]{lewis2015}.
+With the April 2018 data release from Gaia (DR2), we estimate that we will be
+able to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$ 3
+kpc.
 
@@
 less massive stars and earlier for more massive stars.
-Although this effect reduces the usefulness of gyrochronology for stars more
-massive than the Sun, in practise, for Solar-mass stars, down to M dwarfs with
-masses greater than 0.35$M_\odot$, gyrochronology can still be considered an
-effective dating method.
+Further calibration is needed however; a lack of rotation periods and
+reliable ages for old and low-mass stars leaves a question mark hanging over
+the rotational behaviour of stars older than the Sun.
+\TESS\ will provide rotation periods for old stars that fall in the continuous
+viewing zones.
 
-{\bf We will use the rotation periods of TESS field dwarfs to calibrate the
+{\bf We will use the rotation periods of TESS field dwarfs to test the
 gyrochronology relations using galactic kinematics}.
@@
 stars with M$<$ 0.35M$_\odot$.
+
+{\bf We will use the rotation periods of comoving stars identified in the
+first \Gaia\ data release to quantify the accuracy and precision of
+gyrochronology}.
+
+Although this effect reduces the applicability  of gyrochronology for stars
+more massive than the Sun, in practise, for Solar masses and below, down to
+0.35$M_\odot$, gyrochronology can still be considered an effective dating
+method.
 

### 2017-10-04 ms.tex
some big cuts

@@
 \newcommand{\columbia}{1}
+\newcommand{\ww}{2}
+\newcommand{\cca}{3}
+\newcommand{\florida}{4}
+\newcommand{\princeton}{5}
+\newcommand{\nsf}{4}
 \newcommand{\simons}{2}
-\newcommand{\ww}{3}
-\newcommand{\nsf}{4}
-\newcommand{\florida}{5}
-\newcommand{\cca}{6}
 \newcommand{\hubble}{7}
-\newcommand{\princeton}{8}
 
@@
 
-Age is a fundamental stellar parameter of great interest to at galactic
-archaeologists and exoplaneteers alike but unfortunately it is a difficult
-attribute to measure for main sequence F, G, K, and M stars in the field.
+% Age is a fundamental stellar parameter of great interest to at galactic
+% archaeologists and exoplaneteers alike but unfortunately it is a difficult
+% attribute to measure for main sequence F, G, K, and M stars in the field.
 Of the measurable properties for a large ensemble of field stars, rotation
@@
 exoplanet population demographics via gyrochronology.
-
 Angular momentum is carried away though magnetically driven stellar winds,
@@
 amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
-While rotation periods have previously been measured from starspot-induced
-flux modulations for hundreds of stars from the ground, space-based
-photometric surveys have opened the door to homogeneous ensemble measures of
-stellar rotation, and therefore age.
-Using precise light curves available from the TESS mission, we expect to
+Using precise light curves available from the \TESS\ mission, we expect to
 measure the rotation periods of around ... stars.
-Many of these rotation periods will be convertible into ages as they will be
-intermediate age FGK and early M dwarfs.
-Of particular interest will be the expected ~300 gyrochronal ages of exoplanet
-host stars.
-The gyrochronal ages of these planet hosts, combined with their isochronal and
-kinematic ages will allow us to characterize trends in planet properties over
-time and across the galaxy.
+Most of these rotation periods will be convertible into ages as they will be
+intermediate age FGKM dwarfs.
 
-Additionally, as rotation both influences and is influenced {\it by} stellar
-magnetic fields, it is inextricably related to stellar activity.
-With so many \TESS\ targets being M dwarfs which tend to be particularly
-active, understanding the magnetic behavior of these stars has never been more
-important.
+Some outstanding puzzles regarding the nature of magnetic braking remain to be
+answered.
+Firstly, a mysterious gap in the rotation period distribution of \Kepler\
+dwarfs requires either a sharp transition in magnetic dynamo geometry or a gap
+in the local star formation history as an explanation \citep{mcquillan2014,
+davenport2017}.
+Secondly, a transitioning magnetic dynamo appears to be responsible for
+inefficient magnetic braking at old ages in Solar-mass stars.
+{\bf We propose to help answer these questions by providing a catalog of \TESS\
+rotation periods for all FFI and CTL stars.}
+We also intend to use gyrochronology in the cases where it {\it is} applicable
+to learn about the evolution of exoplanets and the galaxy as a whole.
 
-We propose to:
-1. Measure accurate rotation periods for every available TESS FFI and
-two-minute cadence target.
-2. Determine the origin of the mysterious rotation period bimodality
-discovered with Kepler by tracing the rotation period distribution over the
-whole sky, and out to further distances utilizing public Gaia data.
-3. Measure the star formation history across the sky using a new Bayesian
-age-dating system.
-4. Further characterize the intrinsic uncertainty in gyrochronology ages using
-comoving pairs identified in the first Gaia data release.
-5. Infer the distribution of exoplanets as a function of time and galactic
-position.
+% We propose to:
+% 1. Measure accurate rotation periods for every available TESS FFI and
+% two-minute cadence target, beginning with the continuous viewing zones.
+% 2. Determine the origin of the mysterious rotation period bimodality
+% discovered with Kepler by tracing the rotation period distribution over the
+% whole sky, and out to further distances utilizing public Gaia data.
+% 3. Measure the star formation history across the sky using a new Bayesian
+% age-dating system.
+% 4. Further characterize the intrinsic uncertainty in gyrochronology ages using
+% comoving pairs identified in the first Gaia data release.
+% 5. Infer the distribution of exoplanets as a function of time and galactic
+% position.
 
 \section{Scientific Justification}
-\subsection{Rotation period bimodality}
 
-One of the most remarkable results from the Kepler rotation period catalog was
-the discovery of a bimodal period distribution among field stars by
-\citet{mcquillan2013}, which is shown in Figure \ref{fig:davenport}.
-The bimodality separats M dwarfs into two populations, those with rotation
-periods longer than $\sim$ 20 days, and those with periods between $\sim$ 1
-and 20 days.
-Follow-up analysis of the Kepler data by \citet{mcquillan2014} found the
-period bimodality extended to include K dwarfs.
+The bimodal period distribution among field stars discovered by
+\citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
+% The bimodality separates M dwarfs into two populations, those with rotation
+% periods longer than $\sim$ 20 days, and those with periods between $\sim$ 1
+% and 20 days.
 Recently, Co-I Davenport discovered this period bimodality extends throughout
@@
 \ref{fig:davenport} \citep{davenport2017}.
-Two scenarios have been proposed to explain the observed period bimodality.
-In the first scenario, initially proposed by \citet{mcquillan2013}, the
-rotation period distribution reflects the local star formation history, and
-thus the bimodality represents a drop in the star formation rate around 600
-Myr ago.
-This model is supported by both the extension of the bimodality to earlier
-spectral types by \citet{davenport2017}, and also the tentative evidence for
-two distinct proper motion distributions.
-However, such a variation in the star formation rate on short timescales is in
-tension with independent observational efforts to determine the local star
-formation history.
-While color–magnitude diagram inversions from Hipparcos have suggested a
-similarly short timescale variation in star formation of $\sim$ 0.5 Gyr
-\citep{hernandez2000}, other studies find slower variations over several
-Gyr \citep[\eg][]{cignoni2006}.
-Using white dwarf cooling models to infer the local formation history
-(“cosmochronology”) also supports higher star formation several Gyr ago, but
-can rarely achieve age resolution better than $\sim$ 1 Gyr due to small sample
-sizes \citep{tremblay2014}.
-The spatial extent of such coherent and localized variations in star formation
-history is unknown.
-The second scenario explaining this feature is a previously unknown variation
-in the spin-down evolution for low-mass stars.
-In this model, the star formation history would be continuous over the past
-$\sim$ 1 Gyr, and around 600 Myr stars would move quickly through the observed
-period minima due to some phase transition or feedback mechanism.
-While this model is not predicted by angular momentum loss prescriptions,
-rapid transitions in rotation period are observed for stars in young clusters.
-Stars move quickly from the rapidly rotating “convective” sequence (periods of
-$\sim<1$ day) to the “interface” sequence (periods of several days) during the
-first few hundred Myr, with lower mass stars taking longer to make this
-transition as they settle onto the main sequence \citep{barnes2003}.
-Secondly, a gap in chromospheric activity levels for solar-type field stars
-has also been observed \citep{vaughan1980}.
-While this magnetic activity indicator smoothly varies with stellar ages over
-long timescales, the gap indicates a rapid transition phase from “active” to
-“inactive” is present within the first Gyr \citep{pace2009}.
-Thirdly, the angular momentum loss underpinning gyrochronology seems to slow
-for stars older than the Sun, indicating a potential change in the magnetic
-dynamo for slower rotators \citep{angus2015, van-saders2016}.
-
+% \citet{mcquillan2013}, suggests the rotation period distribution reflects the
+% local star formation history, and thus the bimodality represents a drop in the
+% star formation rate around 600 Myr ago.
+% This model is supported by both the extension of the bimodality to earlier
+% spectral types by \citet{davenport2017}, and also the tentative evidence for
+% two distinct proper motion distributions.
+% However, such a variation in the star formation rate on short timescales is in
+% tension with independent observational efforts to determine the local star
+% formation history.
+% While color–magnitude diagram inversions from Hipparcos have suggested a
+% similarly short timescale variation in star formation of $\sim$ 0.5 Gyr
+% \citep{hernandez2000}, other studies find slower variations over several
+% Gyr \citep[\eg][]{cignoni2006}.
+% Using white dwarf cooling models to infer the local formation history
+% (“cosmochronology”) also supports higher star formation several Gyr ago, but
+% can rarely achieve age resolution better than $\sim$ 1 Gyr due to small sample
+% sizes \citep{tremblay2014}.
+% The spatial extent of such coherent and localized variations in star formation
+% history is unknown.
+This feature could either reflect a drop in the star formation rate around 600
+Myr ago or could be explained by a previously unknown variation in the
+spin-down evolution for low-mass stars.
+% In this model, the star formation history would be continuous over the past
+% $\sim$ 1 Gyr, and around 600 Myr stars would move quickly through the observed
+% period minima due to some phase transition or feedback mechanism.
+% While this model is not predicted by angular momentum loss prescriptions,
+% rapid transitions in rotation period are observed for stars in young clusters.
+% Stars move quickly from the rapidly rotating “convective” sequence (periods of
+% $\sim<1$ day) to the “interface” sequence (periods of several days) during the
+% first few hundred Myr, with lower mass stars taking longer to make this
+% transition as they settle onto the main sequence \citep{barnes2003}.
+% Secondly, a gap in chromospheric activity levels for solar-type field stars
+% has also been observed \citep{vaughan1980}.
+% While this magnetic activity indicator smoothly varies with stellar ages over
+% long timescales, the gap indicates a rapid transition phase from “active” to
+% “inactive” is present within the first Gyr \citep{pace2009}.
+% Thirdly, the angular momentum loss underpinning gyrochronology seems to slow
+% for stars older than the Sun, indicating a potential change in the magnetic
+% dynamo for slower rotators \citep{angus2015, van-saders2016}.
 The TESS FFI and CTL targets will provide the ideal dataset to test these two
@@
 with galactic position.
-
-To map the rotation period distribution as a function of galactocentric
-position we will match the TESS FFI and CTL targets to the upcoming data
-release from the Gaia mission \citep{perryman2001}.
-This will also allow us to remove contaminating subgiants and binary stars
-from the sample, leaving only single main sequence stars.
+We will match the TESS FFI and CTL targets to the upcoming data
+release from the Gaia mission \citep{perryman2001} to map the rotation period
+distribution as a function of galactocentric position.
+% This will also allow us to remove contaminating subgiants and binary stars
+% from the sample, leaving only single main sequence stars.
 With the April 2018 data release from Gaia (DR2), we estimate that we will be
@@
 
-\subsection{Gyrochronology}
-
-Gyrochronology is by far the most widely applicable, relatively precise
-age-dating method available.
-Although the phenomenon of magnetic braking has been closely studied since its
-discovery in the mid/late 20th century, the manifestations of magnetic fields
-and their effects on rotation and activity are complex; the physics
-underpinning rotational evolution are far from well-understood.
+% Gyrochronology is by far the most widely applicable, relatively precise
+% age-dating method available.
+% Although the phenomenon of magnetic braking has been closely studied since its
+% discovery in the mid/late 20th century, the manifestations of magnetic fields
+% and their effects on rotation and activity are complex; the physics
+% underpinning rotational evolution are far from well-understood.
 Although the classical spin-down law of \citet{skumanich1972}: Period
@@
 provides an improved fit to the data.
-Once a star reaches a critical Rossby number (the ratio of the rotation period
-to the convective overturn time), magnetic braking efficiency effectively
-drops to zero.
-This means a 5 Gyr star of Solar mass will have the same rotation period as a
-6 Gyr and rotation period cannot be used as an age proxy.
-The transition occurs at around Solar age for Solar mass stars but later for
-less massive stars and earlier for more massive stars.
+% Once a star reaches a critical Rossby number (the ratio of the rotation period
+% to the convective overturn time), magnetic braking efficiency effectively
+% drops to zero.
+% This means a 5 Gyr star of Solar mass will have the same rotation period as a
+% 6 Gyr and rotation period cannot be used as an age proxy.
+% The transition occurs at around Solar age for Solar mass stars but later for
+% less massive stars and earlier for more massive stars.
 Further calibration is needed however; a lack of rotation periods and
@@
 viewing zones.
-
-{\bf We will use the rotation periods of TESS field dwarfs to test the
-gyrochronology relations using galactic kinematics}.
-Many stars in the TESS CTL will also have proper motions, parallaxes,
-positions and radial velocities published in the second \Gaia\ data release.
-These parameters provide the information necessary to calculate both
-galactocentric positions and action angles of the stars, both of which are age
-indicators.
-Vertical action (angular momentum through the plane of the Milky Way)
-increases over time due to orbital heating and has been demonstrated to
-increase with age \citep[\eg][]{west2011}.
-Galactocentric position can also determine a star's age, since different
-stellar populations, thin or thick disk, halo and bulge have different average
-ages.
+Many of these stars will have spectra from the \TESS\/HERMES survey, from
+which it will be possible to derive isochronal ages, enabling further
+calibration of the gyrochronology relations.
+We will also use the rotation periods of TESS field dwarfs over the whole sky
+to test the gyrochronology relations using galactic kinematics.
+Many stars in the TESS CTL will have proper motions, parallaxes, positions and
+radial velocities published in the second \Gaia\ data release.
+These parameters provide the information necessary to calculate galactocentric
+positions and action angles of the stars, both of which are age indicators.
+% Vertical action (angular momentum through the plane of the Milky Way)
+% increases over time due to orbital heating and has been demonstrated to
+% increase with age \citep[\eg][]{west2011}.
+% Galactocentric position can also determine a star's age, since different
+% stellar populations, thin or thick disk, halo and bulge have different average
+% ages.
 PI Angus is currently using the \kepler\ rotation periods of stars in the
@@
 
-{\bf We will use the rotation periods of comoving stars identified in the
-first \Gaia\ data release to quantify the accuracy and precision of
-gyrochronology}.
+In addition, we will use the rotation periods of comoving stars identified in
+the first \Gaia\ data release to quantify the accuracy and precision of
+gyrochronology.
 
-Although this effect reduces the applicability  of gyrochronology for stars
-more massive than the Sun, in practise, for Solar masses and below, down to
-0.35$M_\odot$, gyrochronology can still be considered an effective dating
-method.
+{\bf We will use the rotation periods of planet hosting stars to infer the
+dependence of planet frequency on age and position in the galaxy}.
+\kepler\ has revealed tantalizing hints that exoplanets on short orbital
+periods are more infrequent around young stars \citep{mann2016a, mann2016b,
+rizzuto2017}.
+Additionally, there is growing evidence to suggest that younger stars have
+larger radii.
+Do these trends continue into older ages and do they exist in the field or is
+this a cluster-specific phenomenon?
+We will answer these questions by inferring ages for \TESS\ field stars.
 
-{\bf We will use the rotation periods of planet hosting stars to infer
-the age-dependence of planet frequency.}j Kepler alone has not yet produced a
-sufficient number of stars with rotation periods to reliably characterise the
-time-dependent frequency of exoplanets.
-Mann et al. (2016a), Mann et al. (2016b) and Rizzuto et al. (2017) detected
-transiting exoplanets in four young open clusters observed by the K2 mission.
+\subsection{The age distribution of exoplanets}
 
-\section{Measuring Rotation Periods}
+Of particular interest will be the expected $\sim$ 300 gyrochronal ages of
+exoplanet host stars.
+The gyrochronal ages of these planet hosts, combined with their isochronal and
+kinematic ages may allow us to characterize trends in planet properties over
+time and across the galaxy.
+Additionally, as rotation both influences and is influenced {\it by} stellar
+magnetic fields, it is inextricably related to stellar activity.
+With so many \TESS\ targets being M dwarfs which tend to be particularly
+active, understanding the magnetic behavior of these stars has never been more
+important.
+
+The ages of \TESS\ stars will also aid galactic archaeology studies,
+traditionally performed with red giants, although greater distances can be
+reached with red giants, main sequence stars are far more numerous.
+
+\section{Analysis plan}
+
+\section{Technical Feasibility}
+\subsection{Measuring Rotation Periods}
 In this era of large photometric surveys (Kepler/K2, TESS, WFIRST, LSST, PTF,
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[width=6in, clip=true]{Kalesalad.png}
-\caption{All stars observed during K2 campaign 4, plotted according to their
-equatorial coordinates and colored by their preliminary rotation period.
-These rotation periods were measured using a simple ACF method, applied to
-    everest (Luger et al., 2015) light curves.}
-\label{fig:kalesalad}
-\end{center}
-\end{figure}
-
-We intend to measure rotation periods for all mid to low-mass dwarfs that show
-evidence of rotational modulation in their light curves.
-
-Need to measure rotation periods of all stars if you want to infer the
-distribution of planets as a function of age.
-
-It looks like hotter stars stop spinning down at around the age of the Sun.
-However, that does not necessarily mean that gyrochronology cannot be used to
-infer ages.
-Gyro works for low mass stars down to around 0.35 M$_\odot$ (CITATION), up to
-the age of the Universe.
-It works for Solar-mass stars up to the age of the Sun and for stars more
-massive than the Sun (but less massive than the Kraft break) it works to
-gradually decreasing ages.
-This means that X\% of stars in the TIC are likely to have ages that can be
-determined by gyrochronology.
+% \begin{figure}
+% \begin{center}
+% \includegraphics[width=6in, clip=true]{Kalesalad.png}
+% \caption{All stars observed during K2 campaign 4, plotted according to their
+% equatorial coordinates and colored by their preliminary rotation period.
+% These rotation periods were measured using a simple ACF method, applied to
+%     everest (Luger et al., 2015) light curves.}
+% \label{fig:kalesalad}
+% \end{center}
+% \end{figure}
 

### 2017-10-04 ms.tex
adding pair details, more cuts

@@
 \newcommand{\gaia}{{\it Gaia}}
+\newcommand{\RAVE}{{\it RAVE}}
+\newcommand{\rave}{{\it RAVE}}
 \newcommand{\Teff}{$T_{\mathrm{eff}}$}
@@
 
-% We propose to:
-% 1. Measure accurate rotation periods for every available TESS FFI and
-% two-minute cadence target, beginning with the continuous viewing zones.
-% 2. Determine the origin of the mysterious rotation period bimodality
-% discovered with Kepler by tracing the rotation period distribution over the
-% whole sky, and out to further distances utilizing public Gaia data.
-% 3. Measure the star formation history across the sky using a new Bayesian
-% age-dating system.
-% 4. Further characterize the intrinsic uncertainty in gyrochronology ages using
-% comoving pairs identified in the first Gaia data release.
-% 5. Infer the distribution of exoplanets as a function of time and galactic
-% position.
-
 \section{Scientific Justification}
@@
 \citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
-% The bimodality separates M dwarfs into two populations, those with rotation
-% periods longer than $\sim$ 20 days, and those with periods between $\sim$ 1
-% and 20 days.
 Recently, Co-I Davenport discovered this period bimodality extends throughout
@@
 \ref{fig:davenport} \citep{davenport2017}.
-% \citet{mcquillan2013}, suggests the rotation period distribution reflects the
-% local star formation history, and thus the bimodality represents a drop in the
-% star formation rate around 600 Myr ago.
-% This model is supported by both the extension of the bimodality to earlier
-% spectral types by \citet{davenport2017}, and also the tentative evidence for
-% two distinct proper motion distributions.
-% However, such a variation in the star formation rate on short timescales is in
-% tension with independent observational efforts to determine the local star
-% formation history.
-% While color–magnitude diagram inversions from Hipparcos have suggested a
-% similarly short timescale variation in star formation of $\sim$ 0.5 Gyr
-% \citep{hernandez2000}, other studies find slower variations over several
-% Gyr \citep[\eg][]{cignoni2006}.
-% Using white dwarf cooling models to infer the local formation history
-% (“cosmochronology”) also supports higher star formation several Gyr ago, but
-% can rarely achieve age resolution better than $\sim$ 1 Gyr due to small sample
-% sizes \citep{tremblay2014}.
-% The spatial extent of such coherent and localized variations in star formation
-% history is unknown.
 This feature could either reflect a drop in the star formation rate around 600
@@
 spin-down evolution for low-mass stars.
-% In this model, the star formation history would be continuous over the past
-% $\sim$ 1 Gyr, and around 600 Myr stars would move quickly through the observed
-% period minima due to some phase transition or feedback mechanism.
-% While this model is not predicted by angular momentum loss prescriptions,
-% rapid transitions in rotation period are observed for stars in young clusters.
-% Stars move quickly from the rapidly rotating “convective” sequence (periods of
-% $\sim<1$ day) to the “interface” sequence (periods of several days) during the
-% first few hundred Myr, with lower mass stars taking longer to make this
-% transition as they settle onto the main sequence \citep{barnes2003}.
-% Secondly, a gap in chromospheric activity levels for solar-type field stars
-% has also been observed \citep{vaughan1980}.
-% While this magnetic activity indicator smoothly varies with stellar ages over
-% long timescales, the gap indicates a rapid transition phase from “active” to
-% “inactive” is present within the first Gyr \citep{pace2009}.
-% Thirdly, the angular momentum loss underpinning gyrochronology seems to slow
-% for stars older than the Sun, indicating a potential change in the magnetic
-% dynamo for slower rotators \citep{angus2015, van-saders2016}.
 The TESS FFI and CTL targets will provide the ideal dataset to test these two
@@
 distribution as a function of galactocentric position.
-% This will also allow us to remove contaminating subgiants and binary stars
-% from the sample, leaving only single main sequence stars.
 With the April 2018 data release from Gaia (DR2), we estimate that we will be
@@
 
-% Gyrochronology is by far the most widely applicable, relatively precise
-% age-dating method available.
-% Although the phenomenon of magnetic braking has been closely studied since its
-% discovery in the mid/late 20th century, the manifestations of magnetic fields
-% and their effects on rotation and activity are complex; the physics
-% underpinning rotational evolution are far from well-understood.
 Although the classical spin-down law of \citet{skumanich1972}: Period
@@
 not appear to describe old field stars \citep{angus2015, van-saders2016}.
-Asteroseismic pulsators observed by Kepler, older than the Sun, rotate more
-rapidly than the \citet{skumanich1972} law predicts they should.
-\citet{van-saders2016} find that including a transition to a weakened magnetic
+\citet{van-saders2016} found that including a transition to a weakened magnetic
 braking regime in the gyrochronology models, at a critical Rossby number,
 provides an improved fit to the data.
-% Once a star reaches a critical Rossby number (the ratio of the rotation period
-% to the convective overturn time), magnetic braking efficiency effectively
-% drops to zero.
-% This means a 5 Gyr star of Solar mass will have the same rotation period as a
-% 6 Gyr and rotation period cannot be used as an age proxy.
-% The transition occurs at around Solar age for Solar mass stars but later for
-% less massive stars and earlier for more massive stars.
-Further calibration is needed however; a lack of rotation periods and
-reliable ages for old and low-mass stars leaves a question mark hanging over
-the rotational behaviour of stars older than the Sun.
-\TESS\ will provide rotation periods for old stars that fall in the continuous
-viewing zones.
-Many of these stars will have spectra from the \TESS\/HERMES survey, from
-which it will be possible to derive isochronal ages, enabling further
-calibration of the gyrochronology relations.
+Further calibration is needed however; a lack of rotation periods and reliable
+ages for old and low-mass stars leaves the rotational evolution of stars older
+than the Sun relatively unexplored.
+\TESS\ will provide thousands of new rotation periods for both FFI and
+two-minute cadence stars, with the maximum measurable period strongly
+depending on the length of time it is observed for.
+Of particular interest are the, potentially long, rotation periods measurable
+for stars that fall in the continuous viewing zones.
+Many of these will have spectra from the \TESS\/HERMES survey, from which it
+will be possible to calculate isochronal ages, enabling further calibration of
+the gyrochronology relations.
 We will also use the rotation periods of TESS field dwarfs over the whole sky
-to test the gyrochronology relations using galactic kinematics.
+to test the gyrochronology relations using galactic kinematics with \Gaia\.
 Many stars in the TESS CTL will have proper motions, parallaxes, positions and
@@
 positions and action angles of the stars, both of which are age indicators.
-% Vertical action (angular momentum through the plane of the Milky Way)
-% increases over time due to orbital heating and has been demonstrated to
-% increase with age \citep[\eg][]{west2011}.
-% Galactocentric position can also determine a star's age, since different
-% stellar populations, thin or thick disk, halo and bulge have different average
-% ages.
-PI Angus is currently using the \kepler\ rotation periods of stars in the
-first \Gaia\ data release to calibrate the relations between rotation period
-and vertical action dispersion, two tracers of age.
-Since this relation is likely vary with both galactic longitude and latitude,
-the rotation periods produced from \TESS\ light curves will enable a
-comprehensive, all-sky calibration of these relations.
+% PI Angus is currently using the \kepler\ rotation periods of stars in the
+% first \Gaia\ data release to calibrate the relations between rotation period
+% and vertical action dispersion, two tracers of age.
+% Since this relation is likely vary with both galactic longitude and latitude,
+% the rotation periods produced from \TESS\ light curves will enable a
+% comprehensive, all-sky calibration of these relations.
 Since one of the few ways to accurately age-date fully convective, late M
@@
 In addition, we will use the rotation periods of comoving stars identified in
-the first \Gaia\ data release to quantify the accuracy and precision of
-gyrochronology.
+the first \Gaia\ data release \citep{oh2017} to quantify the accuracy and
+precision of gyrochronology.
+We have identified 38 stars in 19 pairs with \Gaia\ proper motions and radial
+velocities from \RAVE\ that are observable by \TESS, using {\tt tvguide}.
+Of these, 34 stars in 17 pairs are already in the CTL list.
+Because the stars in these pairs are assumed to be recently disrupted
+binaries, we expect them to have the same age and measuring their rotation
+periods will therefore allow us to test gyrochronology.
 
-{\bf We will use the rotation periods of planet hosting stars to infer the
-dependence of planet frequency on age and position in the galaxy}.
-\kepler\ has revealed tantalizing hints that exoplanets on short orbital
-periods are more infrequent around young stars \citep{mann2016a, mann2016b,
-rizzuto2017}.
-Additionally, there is growing evidence to suggest that younger stars have
-larger radii.
-Do these trends continue into older ages and do they exist in the field or is
-this a cluster-specific phenomenon?
+Finally, we hope to use the improved gyrochronology relations, calibrated
+using \TESS\ data to infer the dependence of planet frequency on age and
+position in the galaxy.
+\kepler\ revealed tantalizing hints that exoplanets on short orbital periods
+have larger radii and are more infrequent around young stars \citep{mann2016a,
+mann2016b, rizzuto2017}.
+Do these trends continue into older ages and do they exist in the field?
 We will answer these questions by inferring ages for \TESS\ field stars.
 
-\subsection{The age distribution of exoplanets}
+% \subsection{The age distribution of exoplanets}
 
-Of particular interest will be the expected $\sim$ 300 gyrochronal ages of
-exoplanet host stars.
-The gyrochronal ages of these planet hosts, combined with their isochronal and
-kinematic ages may allow us to characterize trends in planet properties over
-time and across the galaxy.
-Additionally, as rotation both influences and is influenced {\it by} stellar
-magnetic fields, it is inextricably related to stellar activity.
-With so many \TESS\ targets being M dwarfs which tend to be particularly
-active, understanding the magnetic behavior of these stars has never been more
-important.
+% Of particular interest will be the expected $\sim$ 300 gyrochronal ages of
+% exoplanet host stars.
+% The gyrochronal ages of these planet hosts, combined with their isochronal and
+% kinematic ages may allow us to characterize trends in planet properties over
+% time and across the galaxy.
+% Additionally, as rotation both influences and is influenced {\it by} stellar
+% magnetic fields, it is inextricably related to stellar activity.
+% With so many \TESS\ targets being M dwarfs which tend to be particularly
+% active, understanding the magnetic behavior of these stars has never been more
+% important.
 
-The ages of \TESS\ stars will also aid galactic archaeology studies,
-traditionally performed with red giants, although greater distances can be
-reached with red giants, main sequence stars are far more numerous.
-
-\section{Analysis plan}
+% The ages of \TESS\ stars will also aid galactic archaeology studies,
+% traditionally performed with red giants, although greater distances can be
+% reached with red giants, main sequence stars are far more numerous.
 
 \section{Technical Feasibility}
-\subsection{Measuring Rotation Periods}
-In this era of large photometric surveys (Kepler/K2, TESS, WFIRST, LSST, PTF,
-PanStarrs and more) rotation periods are quickly becoming one of the most
-accessible properties of stars.
-Precise light curves produced by these surveys often reveal the presence of
-dark spots on the surfaces of cool stars which revolve with the stellar
-surface creating an overall dimming effect once every rotation period.
-Dark spotted regions and bright faculae leave a characteristic trace in the
-light curve from which a rotation period can be inferred.
-The extraction of a rotation period from a light curve can be as
-straight-forward as computing a Lomb-Scargle periodogram or, in cases where
-the signal is less clear, can be inferred via modelling the correlation
-between data points.
+The extraction of a rotation period from a light curve can be as simple as
+computing a Lomb-Scargle periodogram or, in cases where the signal is less
+clear, can be inferred via modeling the correlation between data points.
 This latter approach could involve either computing an autocorrelation
-function (McQuillan et al., 2013), or fitting a Gaussian process to the time
-series (Angus et al., 2017, Foreman-Mackey et al., 2017).
-Signals produced by the rotation of spotted stars can have amplitudes of a few
-percent of the total flux, but can also have very low amplitudes of a few
-parts per million.
-The frequencies of high amplitude signals are easy to measure and, in these
-cases, most measured frequencies will agree regardless of the technique used
-to measure them.
-Similarly, short-period signals are easy to measure, especially if light curve
-variations are sinusoidal in shape.
-In the low amplitude and long period cases however, care is needed to separate
-real astrophysical signal from instrumental systematics.
-
+function \citep{mcquillan2013}, or fitting a Gaussian process to the time
+series \citep{angus2017, foreman-mackey2017}.
+In cases where signals have long periods and low amplitudes, care is needed to
+separate real astrophysical signal from instrumental systematics.
 The measurement of rotation periods is less sensitive to crowding and source
@@
 datasets.
-
 The precision of existing photometric de-trending methods degrades in crowded

### 2017-10-04 ms.tex
adding some citations

@@
 In addition, we will use the rotation periods of comoving stars identified in
-the first \Gaia\ data release \citep{oh2017} to quantify the accuracy and
+the first \Gaia\ data release \citep{oh2016} to quantify the accuracy and
 precision of gyrochronology.
@@
 \kepler\ revealed tantalizing hints that exoplanets on short orbital periods
-have larger radii and are more infrequent around young stars \citep{mann2016a,
-mann2016b, rizzuto2017}.
+have larger radii and are more infrequent around young stars \citep{mann2017a,
+mann2017b, rizzuto2017}.
 Do these trends continue into older ages and do they exist in the field?

### 2017-10-04 ms.tex
removing the authors

@@
 \newcommand{\rave}{{\it RAVE}}
+\newcommand{\hermes}{{\it HERMES}}
+\newcommand{\HERMES}{{\it HERMES}}
 \newcommand{\Teff}{$T_{\mathrm{eff}}$}
@@
 
-\author{R. Angus\altaffilmark{\columbia}
-        J. R. Davenport\altaffilmark{\ww}
-        D. Foreman-Mackey\altaffilmark{\cca},
-        S. Oh\altaffilmark{\princeton},
-        D. Busazi\altaffilmark{\florida},
-        A. Mann\altaffilmark{\columbia}
-        D. Kipping\altaffilmark{\columbia}
-}
+% \author{R. Angus\altaffilmark{\columbia}
+%         J. R. Davenport\altaffilmark{\ww}
+%         D. Foreman-Mackey\altaffilmark{\cca},
+%         S. Oh\altaffilmark{\princeton},
+%         D. Busazi\altaffilmark{\florida},
+%         A. Mann\altaffilmark{\columbia}
+%         D. Kipping\altaffilmark{\columbia}
+% }
 
-\altaffiltext{\columbia}{Department of Astronomy, Columbia University, New
-York, NY}
-\altaffiltext{\simons}{Simons Fellow, RuthAngus@gmail.com}
-\altaffiltext{\ww}{Western Washington University, Bellingham, WA}
-\altaffiltext{\nsf}{NSF Fellow}
-\altaffiltext{\florida}{Florida Gulf Coast University, Fort Myers, FL}
-\altaffiltext{\cca}{Center for Computational Astronomy, Flatiron Institute,
-New York, NY}
-\altaffiltext{\hubble}{Hubble Fellow}
-\altaffiltext{\princeton}{Princeton University, Princeton, NJ}
+% \altaffiltext{\columbia}{Department of Astronomy, Columbia University, New
+% York, NY}
+% \altaffiltext{\simons}{Simons Fellow, RuthAngus@gmail.com}
+% \altaffiltext{\ww}{Western Washington University, Bellingham, WA}
+% \altaffiltext{\nsf}{NSF Fellow}
+% \altaffiltext{\florida}{Florida Gulf Coast University, Fort Myers, FL}
+% \altaffiltext{\cca}{Center for Computational Astronomy, Flatiron Institute,
+% New York, NY}
+% \altaffiltext{\hubble}{Hubble Fellow}
+% \altaffiltext{\princeton}{Princeton University, Princeton, NJ}
 
@@
 measure the rotation periods of around ... stars.
-Most of these rotation periods will be convertible into ages as they will be
-intermediate age FGKM dwarfs.
+For many of these stars it will be possible to infer an age from their
+rotation periods via gyrochronology.
 
-Some outstanding puzzles regarding the nature of magnetic braking remain to be
-answered.
+Some outstanding mysteries regarding the nature of magnetic braking have been
+revealed, but thus-far unanswered by, \kepler.
 Firstly, a mysterious gap in the rotation period distribution of \Kepler\
@@
 We also intend to use gyrochronology in the cases where it {\it is} applicable
-to learn about the evolution of exoplanets and the galaxy as a whole.
+to learn about the evolution of exoplanets across the galaxy.
 
@@
 Recently, Co-I Davenport discovered this period bimodality extends throughout
-all masses in the Kepler rotation sample for nearby stars, as shown in Figure
-\ref{fig:davenport} \citep{davenport2017}.
+all masses in the Kepler rotation sample for nearby stars
+\citep{davenport2017}.
 This feature could either reflect a drop in the star formation rate around 600
@@
 scenarios explaining the appearance of a period bimodality.
-If the bimodal period distribution reflects a age discontinuous age
-distribution, the feature should be local; it could
-disappear at greater distances or along different lines of sight.
+If the bimodal period distribution reflects a discontinuous age distribution,
+the feature should be local and may disappear at greater distances or along
+different lines of sight.
 However, if the bimodality is truly due to a transition point in the spin-down
@@
 depending on the length of time it is observed for.
-Of particular interest are the, potentially long, rotation periods measurable
+Of particular interest are the potentially long rotation periods measurable
 for stars that fall in the continuous viewing zones.
-Many of these will have spectra from the \TESS\/HERMES survey, from which it
+Many of these will have spectra from the \TESS\ /\HERMES\ survey, from which it
 will be possible to calculate isochronal ages, enabling further calibration of
@@
 We will also use the rotation periods of TESS field dwarfs over the whole sky
-to test the gyrochronology relations using galactic kinematics with \Gaia\.
+to test the gyrochronology relations using galactic kinematics.
 Many stars in the TESS CTL will have proper motions, parallaxes, positions and
@@
 dwarfs is via kinematics, these new relations will help to infer ages for all
-stars with M$<$ 0.35M$_\odot$.
+stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
 
@@
 binaries, we expect them to have the same age and measuring their rotation
-periods will therefore allow us to test gyrochronology.
+periods will therefore provide a test of gyrochronology.
 
@@
 Do these trends continue into older ages and do they exist in the field?
-We will answer these questions by inferring ages for \TESS\ field stars.
+We will answer these questions by measuring rotation periods for all \TESS\
+field stars.
 
@@
 \section{Expected Impact}
-We will provide light curves and rotation periods for both the two-minute
-cadence and FFI targets.
+We will provide rotation periods for all two-minute cadence and FFI targets.
 

### 2017-10-04 ms.tex

@@
-\documentclass[useAMS, usenatbib, preprint, 12pt]{aastex}
-\usepackage{cite, natbib}
-\usepackage{float}
-\usepackage{epsfig}
-\usepackage{cases}
-\usepackage[section]{placeins}
-\usepackage{graphicx, subfigure}
-\usepackage{color}
-\usepackage{bm}
+\documentclass[letterpaper,12pt,preprint]{hack_aastex}
+
+\input{dfm_stylez}
+\pagestyle{myheadings}
+\markright{\textsf{\footnotesize %
+    Investigating magnetic dynamo evolution with TESS field dwarfs}}
+
+%\documentclass[useAMS, usenatbib, preprint, 12pt]{aastex}
+%\usepackage{cite, natbib}
+%\usepackage{float}
+%\usepackage{epsfig}
+%\usepackage{cases}
+%\usepackage[section]{placeins}
+%\usepackage{graphicx, subfigure}
+%\usepackage{color}
+%\usepackage{bm}
 
@@
 
-\section{Introduction}
+%\paragraph{Introduction}
 
@@
 Angular momentum is carried away though magnetically driven stellar winds,
-which slows the star’s rotation over cosmic time.
+which slows the star's rotation over cosmic time.
 This rotation-based `clock' is known as gyrochronology.
-Cool spots on the star’s surface rotate in-to and out-of view, creating small
+Cool spots on the star's surface rotate in-to and out-of view, creating small
 amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
@@
 
-\section{Scientific Justification}
+\paragraph{Scientific Justification}
 
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics[width=3in, clip=true]{Davenport.png}
-\caption{From Davenport (2017).
+\ssfigure{Davenport.png}{0.5}{%
+From Davenport (2017).
     Rotation period as a function of effective
@@
 rotation period bimodality is revealed to exist across all temperatures shown.
-    The red line is a 600 Myr rotational isochrone (also known as a gyrochrone).}
+    The red line is a 600 Myr rotational isochrone (also known as a gyrochrone).
 \label{fig:davenport}
-\end{center}
-\end{figure}
+}
+%\begin{figure}
+%\begin{center}
+%\includegraphics[width=3in, clip=true]{Davenport.png}
+%\caption{From Davenport (2017).
+%    Rotation period as a function of effective
+%temperature for the full McQuillan et al., (2014) Kepler sample in black, and
+%the subset of these stars that also feature in the TGAS Gaia DR1 catalogue in
+%blue. Contaminating giants have been removed from the blue sample and the
+%rotation period bimodality is revealed to exist across all temperatures shown.
+%    The red line is a 600 Myr rotational isochrone (also known as a gyrochrone).}
+%\label{fig:davenport}
+%\end{center}
+%\end{figure}
 
@@
 
-\section{Technical Feasibility}
+\paragraph{Technical Feasibility}
 The extraction of a rotation period from a light curve can be as simple as
@@
 
-\section{Expected Impact}
+\paragraph{Expected Impact}
 We will provide rotation periods for all two-minute cadence and FFI targets.
 
-\section{Budget Justification}
+\paragraph{Budget Justification}
 PI Angus intends to use the budget to employ a student for 3-4 months.
@@
 
+\paragraph{References}
+
+{\footnotesize
 \bibliography{references}
+}
 

### 2017-10-04 ms.tex
adding priority list

@@
 \documentclass[letterpaper,12pt,preprint]{hack_aastex}
+
+% Still to do: estimate the number of 16th mag stars and brighter in the whole
+% sky.
+% Estimate computation power available.
+% Come up with a better target and priorities list.
 
@@
 \markright{\textsf{\footnotesize %
-    Investigating magnetic dynamo evolution with TESS field dwarfs}}
-
-%\documentclass[useAMS, usenatbib, preprint, 12pt]{aastex}
-%\usepackage{cite, natbib}
-%\usepackage{float}
-%\usepackage{epsfig}
-%\usepackage{cases}
-%\usepackage[section]{placeins}
-%\usepackage{graphicx, subfigure}
-%\usepackage{color}
-%\usepackage{bm}
+    Investigating rotational evolution with TESS field dwarfs}}
 
@@
 \newcommand{\TESS}{{\it TESS}}
+\newcommand{\tess}{{\it TESS}}
 \newcommand{\LSST}{{\it LSST}}
@@
 
-%\paragraph{Introduction}
-
-% Age is a fundamental stellar parameter of great interest to at galactic
-% archaeologists and exoplaneteers alike but unfortunately it is a difficult
-% attribute to measure for main sequence F, G, K, and M stars in the field.
 Of the measurable properties for a large ensemble of field stars, rotation
@@
 amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
-Using precise light curves available from the \TESS\ mission, we expect to
-measure the rotation periods of around ... stars.
+Based on results from \Kepler, using precise light curves available from the
+\TESS\ mission, we expect to measure the rotation periods of around 20\% of FG
+stars and 80\% of KM stars in the Candidate Target List (CTL)\footnote{The
+difference between these percentages comes from a dramatic increase in
+magnetic activity towards lower masses.}.
 For many of these stars it will be possible to infer an age from their
@@
 {\bf We propose to help answer these questions by providing a catalog of \TESS\
-rotation periods for all FFI and CTL stars.}
+rotation periods for FFI and CTL stars.}
 We also intend to use gyrochronology in the cases where it {\it is} applicable
@@
 distribution as a function of galactocentric position.
+For the all-sky analysis we will focus on F and G stars since the 27.5 day
+baseline of \TESS\ fields will limit measureable rotation periods to less than
+around 15 days.
 With the April 2018 data release from Gaia (DR2), we estimate that we will be
-able to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$ 3
+able to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$3
 kpc.
+To characterize the rotation period bimodality as a function of galactic
+position, we propose to analyse the light curves of 10,000 FFI and two-minute
+targets that overlap with the Tycho-Gaia Astrometric Solution (TGAS) catalog,
+selecting FGKM dwarfs with a range of galactic positions.
+The FFI targets will be essential in order to reach $\sim$3kpc distances.
+Of these, the stars lying in the CVZs will be our highest priority.
+M dwarfs rotation periods in the CVZz will allow us to study the rotation
+period bimodality at lower masses, since the bimodality appears at longer
+rotation periods for lower mass stars.
 
@@
 }
-%\begin{figure}
-%\begin{center}
-%\includegraphics[width=3in, clip=true]{Davenport.png}
-%\caption{From Davenport (2017).
-%    Rotation period as a function of effective
-%temperature for the full McQuillan et al., (2014) Kepler sample in black, and
-%the subset of these stars that also feature in the TGAS Gaia DR1 catalogue in
-%blue. Contaminating giants have been removed from the blue sample and the
-%rotation period bimodality is revealed to exist across all temperatures shown.
-%    The red line is a 600 Myr rotational isochrone (also known as a gyrochrone).}
-%\label{fig:davenport}
-%\end{center}
-%\end{figure}
 
@@
 Of particular interest are the potentially long rotation periods measurable
-for stars that fall in the continuous viewing zones.
+for stars that fall in the Continuous Viewing Zones (CVZs).
 Many of these will have spectra from the \TESS\ /\HERMES\ survey, from which it
@@
 stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
+In order to calibrate gyrochronology using rotation periods of FGK dwarfs with
+\tess\ / \hermes\ spectra we propose to prioritize rotation periods of stars
+in the CVZs with spectra, followed by those without.
 
@@
 field stars.
-
-% \subsection{The age distribution of exoplanets}
-
-% Of particular interest will be the expected $\sim$ 300 gyrochronal ages of
-% exoplanet host stars.
-% The gyrochronal ages of these planet hosts, combined with their isochronal and
-% kinematic ages may allow us to characterize trends in planet properties over
-% time and across the galaxy.
-% Additionally, as rotation both influences and is influenced {\it by} stellar
-% magnetic fields, it is inextricably related to stellar activity.
-% With so many \TESS\ targets being M dwarfs which tend to be particularly
-% active, understanding the magnetic behavior of these stars has never been more
-% important.
-
-% The ages of \TESS\ stars will also aid galactic archaeology studies,
-% traditionally performed with red giants, although greater distances can be
-% reached with red giants, main sequence stars are far more numerous.
 
@@
 
-% \begin{figure}
-% \begin{center}
-% \includegraphics[width=6in, clip=true]{Kalesalad.png}
-% \caption{All stars observed during K2 campaign 4, plotted according to their
-% equatorial coordinates and colored by their preliminary rotation period.
-% These rotation periods were measured using a simple ACF method, applied to
-%     everest (Luger et al., 2015) light curves.}
-% \label{fig:kalesalad}
-% \end{center}
-% \end{figure}
+The full set of available \TESS\ FFI and two-minute cadence light curves will
+be exceedingly large and we intend to prioritize certain groups of stars for
+period analysis, calculating a simple (and computationally fast) Lomb-Scargle
+periodogram for the low priority objects and running a full probabilistic
+period analysis on the high priority targets.
+High priority targets include: comoving pairs, stars in the CVZs (especially
+those with spectra), stars in TGAS, stars at large distances, any potential
+asteroseismic dwarfs and exoplanet hosts.
+% Of these, the stars that also lie in the CVZs as well as the identified \Gaia\
+% comoving pairs will be our highest priority.
+% Our next priority will be stars in the CVZs, with spectra, that do not appear
+% in TGAS, followed by those without spectra.
+% M dwarfs rotation periods in the CVZz will allow us to study the rotation
+% period bimodality at lower masses, since the bimodality appears at longer
+% rotation periods for lower mass stars.
+% We will then focus on measuring rotation periods for a sample of stars that is
+% representative of the full set of FFI and CTL targets in order to enable
+% exoplanet population studies.
+An large amount of computational resources are available to us, with access to
+the Columbia supercomputer as well as facilities at the Flatiron Institute.
+% We expect that each target will require around one minute of computation time
+% and have access to around 10,000 computing hours.
+% We therefore expect to analyze around 60,000 light curves.
+
+We have already begun to construct a rotation period catalog of \ktwo\ stars,
+for which we have received funding from the Archival Data Analysis Program
+(ADAP).
+Figure \ref{fig:kalesalad} shows a map of preliminary rotation periods for one
+of four \ktwo\ fields processed so far.
+
+\ssfigure{Kalesalad.png}{0.5}{%
+All stars observed during K2 campaign 4, plotted according to their
+equatorial coordinates and colored by their preliminary rotation period.
+These rotation periods were measured using a simple ACF method, applied to
+    everest (Luger et al., 2015) light curves.
+\label{fig:kalesalad}
+}
 
 \paragraph{Expected Impact}
-We will provide rotation periods for all two-minute cadence and FFI targets.
+We will provide rotation period posterior probability density functions and
+best fit estimates for around 30,000 two-minute cadence and FFI targets.
 

### 2017-10-05 ms.tex
removing exoplanet stuffs

@@
 
-% \author{R. Angus\altaffilmark{\columbia}
-%         J. R. Davenport\altaffilmark{\ww}
-%         D. Foreman-Mackey\altaffilmark{\cca},
-%         S. Oh\altaffilmark{\princeton},
-%         D. Busazi\altaffilmark{\florida},
-%         A. Mann\altaffilmark{\columbia}
-%         D. Kipping\altaffilmark{\columbia}
-% }
-
-% \altaffiltext{\columbia}{Department of Astronomy, Columbia University, New
-% York, NY}
-% \altaffiltext{\simons}{Simons Fellow, RuthAngus@gmail.com}
-% \altaffiltext{\ww}{Western Washington University, Bellingham, WA}
-% \altaffiltext{\nsf}{NSF Fellow}
-% \altaffiltext{\florida}{Florida Gulf Coast University, Fort Myers, FL}
-% \altaffiltext{\cca}{Center for Computational Astronomy, Flatiron Institute,
-% New York, NY}
-% \altaffiltext{\hubble}{Hubble Fellow}
-% \altaffiltext{\princeton}{Princeton University, Princeton, NJ}
-
 Of the measurable properties for a large ensemble of field stars, rotation
@@
 rotation periods for FFI and CTL stars.}
-We also intend to use gyrochronology in the cases where it {\it is} applicable
-to learn about the evolution of exoplanets across the galaxy.
+% We also intend to use gyrochronology in the cases where it {\it is} applicable
+% to learn about the evolution of exoplanets across the galaxy.
 
@@
 
-Finally, we hope to use the improved gyrochronology relations, calibrated
-using \TESS\ data to infer the dependence of planet frequency on age and
-position in the galaxy.
-\kepler\ revealed tantalizing hints that exoplanets on short orbital periods
-have larger radii and are more infrequent around young stars \citep{mann2017a,
-mann2017b, rizzuto2017}.
-Do these trends continue into older ages and do they exist in the field?
-We will answer these questions by measuring rotation periods for all \TESS\
-field stars.
+% Finally, we hope to use the improved gyrochronology relations, calibrated
+% using \TESS\ data to infer the dependence of planet frequency on age and
+% position in the galaxy.
+% \kepler\ revealed tantalizing hints that exoplanets on short orbital periods
+% have larger radii and are more infrequent around young stars \citep{mann2017a,
+% mann2017b, rizzuto2017}.
+% Do these trends continue into older ages and do they exist in the field?
+% We will answer these questions by measuring rotation periods for all \TESS\
+% field stars.
 
@@
 % exoplanet population studies.
-An large amount of computational resources are available to us, with access to
-the Columbia supercomputer as well as facilities at the Flatiron Institute.
+% An large amount of computational resources are available to us, with access to
+% the Columbia supercomputer as well as facilities at the Flatiron Institute.
 % We expect that each target will require around one minute of computation time

### 2017-10-05 ms.tex
clearing up

@@
 The FFI targets will be essential in order to reach $\sim$3kpc distances.
-Of these, the stars lying in the CVZs will be our highest priority.
+% Of these, the stars lying in the CVZs will be our highest priority.
 M dwarfs rotation periods in the CVZz will allow us to study the rotation
@@
 rotation periods for lower mass stars.
+Unlike \kepler\ and \ktwo\ light curves, \TESS\ provides the all-sky coverage
+necessary to map the period bimodality across the galaxy.
 
@@
 for stars that fall in the Continuous Viewing Zones (CVZs).
-Many of these will have spectra from the \TESS\ /\HERMES\ survey, from which it
+Many of these will have spectra from the \TESS --\HERMES\ survey, from which it
 will be possible to calculate isochronal ages, enabling further calibration of
@@
 to test the gyrochronology relations using galactic kinematics.
-Many stars in the TESS CTL will have proper motions, parallaxes, positions and
-radial velocities published in the second \Gaia\ data release.
+Many stars observable by \tess\ will have proper motions, parallaxes,
+positions and radial velocities published in the second \Gaia\ data release.
 These parameters provide the information necessary to calculate galactocentric
@@
 stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
-In order to calibrate gyrochronology using rotation periods of FGK dwarfs with
-\tess\ / \hermes\ spectra we propose to prioritize rotation periods of stars
-in the CVZs with spectra, followed by those without.
+% In order to calibrate gyrochronology using rotation periods of FGK dwarfs with
+% \tess--\hermes\ spectra we propose to prioritize rotation periods of stars
+% in the CVZs with spectra, followed by those without.
 

### 2017-10-05 ms.tex
expanding impact

@@
 High priority targets include: comoving pairs, stars in the CVZs (especially
-those with spectra), stars in TGAS, stars at large distances, any potential
-asteroseismic dwarfs and exoplanet hosts.
+those with spectra), stars in TGAS, stars at large distances, plus any
+potential asteroseismic dwarfs and exoplanet hosts.
 % Of these, the stars that also lie in the CVZs as well as the identified \Gaia\
@@
 \paragraph{Expected Impact}
-We will provide rotation period posterior probability density functions and
-best fit estimates for around 30,000 two-minute cadence and FFI targets.
+Initially we will provide rotation period posterior probability density
+functions and best fit estimates for around 30,000 high priority two-minute
+cadence and FFI targets.
+We will then produce a catalog of rotation periods for lower priority stars
+using a combination of Lomb-Scargle periodogram and autocorrelation
+techniques.
 

### 2017-10-05 ms.tex
mag limit

@@
 inefficient magnetic braking at old ages in Solar-mass stars.
-{\bf We propose to help answer these questions by providing a catalog of \TESS\
-rotation periods for FFI and CTL stars.}
+{\bf We propose to help answer these questions by providing a catalog of
+rotation periods for \TESS\ Full Frame Image (FFI) stars.}
 % We also intend to use gyrochronology in the cases where it {\it is} applicable
@@
 potential asteroseismic dwarfs and exoplanet hosts.
+Based on the expected photometric precision of \tess\ \citep{sullivan2015}, we
+expect to be able to measure rotation periods for FFI stars down to 16th
+magnitude.
+We note, however that not all stars will have measurable rotation periods,
+even if they are bright enough --- not all stars show rotational variability
+in their light curves either because they are pole-on or because they are
+inactive.
 % Of these, the stars that also lie in the CVZs as well as the identified \Gaia\
@@
 techniques.
+Finally, we will release a catalog of rotation periods for two-minute cadence
+targets.
+We will also release an open source python package for measuring rotation
+periods of both FFI and two-minute cadence targets.
 
@@
 gyrochronology of comoving pairs.
-
+The remaining budget will be used for traveling to meet with other Co-Is in
+order to perform this analysis.
 \paragraph{References}

### 2017-10-05 ms.tex
focussing on FFI stars

@@
 \markright{\textsf{\footnotesize %
-    Investigating rotational evolution with TESS field dwarfs}}
+    Investigating rotational evolution with TESS FFI field dwarfs}}
 
@@
 spin-down evolution for low-mass stars.
-The TESS FFI and CTL targets will provide the ideal dataset to test these two
+The TESS FFI targets will provide the ideal dataset to test these two
 scenarios explaining the appearance of a period bimodality.
@@
 with galactic position.
-We will match the TESS FFI and CTL targets to the upcoming data
+We will match the TESS FFI targets to the upcoming data
 release from the Gaia mission \citep{perryman2001} to map the rotation period
@@
 positions and action angles of the stars, both of which are age indicators.
-% PI Angus is currently using the \kepler\ rotation periods of stars in the
-% first \Gaia\ data release to calibrate the relations between rotation period
-% and vertical action dispersion, two tracers of age.
-% Since this relation is likely vary with both galactic longitude and latitude,
-% the rotation periods produced from \TESS\ light curves will enable a
-% comprehensive, all-sky calibration of these relations.
 Since one of the few ways to accurately age-date fully convective, late M
@@
 stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
-% In order to calibrate gyrochronology using rotation periods of FGK dwarfs with
-% \tess--\hermes\ spectra we propose to prioritize rotation periods of stars
-% in the CVZs with spectra, followed by those without.
 
@@
 velocities from \RAVE\ that are observable by \TESS, using {\tt tvguide}.
-Of these, 34 stars in 17 pairs are already in the CTL list.
-Because the stars in these pairs are assumed to be recently disrupted
-binaries, we expect them to have the same age and measuring their rotation
-periods will therefore provide a test of gyrochronology.
-
-% Finally, we hope to use the improved gyrochronology relations, calibrated
-% using \TESS\ data to infer the dependence of planet frequency on age and
-% position in the galaxy.
-% \kepler\ revealed tantalizing hints that exoplanets on short orbital periods
-% have larger radii and are more infrequent around young stars \citep{mann2017a,
-% mann2017b, rizzuto2017}.
-% Do these trends continue into older ages and do they exist in the field?
-% We will answer these questions by measuring rotation periods for all \TESS\
-% field stars.
+Of these, 34 stars in 17 pairs are in the CTL list.
+Since these stars would make interesting targets for both asteroseismology and
+exoplanet search, we include a list of them with this proposal.
+The stars in these pairs are assumed to be recently disrupted binaries so we
+expect them to have the same age; measuring their rotation periods will
+therefore provide a test of gyrochronology.
 
@@
 
-The full set of available \TESS\ FFI and two-minute cadence light curves will
-be exceedingly large and we intend to prioritize certain groups of stars for
-period analysis, calculating a simple (and computationally fast) Lomb-Scargle
-periodogram for the low priority objects and running a full probabilistic
-period analysis on the high priority targets.
+The full set of available \TESS\ FFI light curves will be exceedingly large
+and we intend to prioritize certain groups of stars for period analysis,
+calculating a simple (and computationally fast) Lomb-Scargle periodogram for
+the low priority objects and running a full probabilistic period analysis on
+the high priority targets.
 High priority targets include: comoving pairs, stars in the CVZs (especially
@@
 inactive.
-% Of these, the stars that also lie in the CVZs as well as the identified \Gaia\
-% comoving pairs will be our highest priority.
-% Our next priority will be stars in the CVZs, with spectra, that do not appear
-% in TGAS, followed by those without spectra.
-% M dwarfs rotation periods in the CVZz will allow us to study the rotation
-% period bimodality at lower masses, since the bimodality appears at longer
-% rotation periods for lower mass stars.
-% We will then focus on measuring rotation periods for a sample of stars that is
-% representative of the full set of FFI and CTL targets in order to enable
-% exoplanet population studies.
-% An large amount of computational resources are available to us, with access to
-% the Columbia supercomputer as well as facilities at the Flatiron Institute.
-% We expect that each target will require around one minute of computation time
-% and have access to around 10,000 computing hours.
-% We therefore expect to analyze around 60,000 light curves.
 
@@
 \paragraph{Expected Impact}
-Initially we will provide rotation period posterior probability density
-functions and best fit estimates for around 30,000 high priority two-minute
-cadence and FFI targets.
-We will then produce a catalog of rotation periods for lower priority stars
-using a combination of Lomb-Scargle periodogram and autocorrelation
+Initially, we will provide rotation period posterior probability density
+functions and best fit estimates for around 30,000 high priority FFI targets.
+We will then produce a catalog of rotation periods for lower priority FFI
+stars using a combination of Lomb-Scargle periodogram and autocorrelation
 techniques.
@@
 targets.
+The full catalog will likely contain several million rotation periods.
 We will also release an open source python package for measuring rotation

### 2017-10-05 ms.tex
small edits

@@
 \TESS\ mission, we expect to measure the rotation periods of around 20\% of FG
-stars and 80\% of KM stars in the Candidate Target List (CTL)\footnote{The
+stars and 80\% of KM stars in the \tess\ Full Frame Images (FFIs)\footnote{The
 difference between these percentages comes from a dramatic increase in
-magnetic activity towards lower masses.}.
+magnetic activity, and therefore more easily measured rotation periods,
+towards lower masses.}.
 For many of these stars it will be possible to infer an age from their
 rotation periods via gyrochronology.
-
-Some outstanding mysteries regarding the nature of magnetic braking have been
-revealed, but thus-far unanswered by, \kepler.
+Some outstanding mysteries regarding the nature of magnetic braking were
+revealed, but thus-far unanswered, by \kepler.
 Firstly, a mysterious gap in the rotation period distribution of \Kepler\
@@
 inefficient magnetic braking at old ages in Solar-mass stars.
-{\bf We propose to help answer these questions by providing a catalog of
+{\bf We propose to answer these questions by providing a catalog of
 rotation periods for \TESS\ Full Frame Image (FFI) stars.}
@@
 \citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
-Recently, Co-I Davenport discovered this period bimodality extends throughout
-all masses in the Kepler rotation sample for nearby stars
-\citep{davenport2017}.
+Recently, Co-I Davenport found this period bimodality extends throughout all
+masses in the Kepler rotation sample for nearby stars \citep{davenport2017}.
 This feature could either reflect a drop in the star formation rate around 600
@@
 with galactic position.
-We will match the TESS FFI targets to the upcoming data
+We will match the TESS FFI targets to the upcoming second data
 release from the Gaia mission \citep{perryman2001} to map the rotation period
@@
 around 15 days.
-With the April 2018 data release from Gaia (DR2), we estimate that we will be
-able to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$3
-kpc.
-To characterize the rotation period bimodality as a function of galactic
-position, we propose to analyse the light curves of 10,000 FFI and two-minute
-targets that overlap with the Tycho-Gaia Astrometric Solution (TGAS) catalog,
-selecting FGKM dwarfs with a range of galactic positions.
-The FFI targets will be essential in order to reach $\sim$3kpc distances.
+% With the April 2018 data release from Gaia (DR2), we estimate that we will be
+% able to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$3
+% kpc.
+% To characterize the rotation period bimodality as a function of galactic
+% position, we propose to analyse the light curves of 10,000 FFI and two-minute
+% targets that overlap with the Tycho-Gaia Astrometric Solution (TGAS) catalog,
+% selecting FGKM dwarfs with a range of galactic positions.
+% The FFI targets will be essential in order to reach $\sim$3kpc distances.
 % Of these, the stars lying in the CVZs will be our highest priority.
-M dwarfs rotation periods in the CVZz will allow us to study the rotation
-period bimodality at lower masses, since the bimodality appears at longer
-rotation periods for lower mass stars.
-Unlike \kepler\ and \ktwo\ light curves, \TESS\ provides the all-sky coverage
-necessary to map the period bimodality across the galaxy.
+The longer time coverage in the CVZz will allow us to study the rotation
+period bimodality in K and M dwarfs, since the bimodality appears at longer
+rotation periods for these lower mass stars.
+{\bf Unlike \kepler\ and \ktwo\ light curves, \TESS\ provides the all-sky
+coverage necessary to map the period bimodality across the galaxy.}
 
@@
 
-Although the classical spin-down law of \citet{skumanich1972}: Period
-$\propto$ Age$^1/2$ holds for all open clusters with measured periods, it does
+Although the classical spin-down law of \citet{skumanich1972} (Period
+$\propto$ Age$^1/2$) holds for all open clusters with measured periods, it does
 not appear to describe old field stars \citep{angus2015, van-saders2016}.
@@
 than the Sun relatively unexplored.
-\TESS\ will provide thousands of new rotation periods for both FFI and
-two-minute cadence stars, with the maximum measurable period strongly
-depending on the length of time it is observed for.
+\TESS\ will provide thousands of new rotation periods, with the maximum
+measurable period strongly depending on the length of time it is observed for.
 Of particular interest are the potentially long rotation periods measurable
@@
 PI Angus intends to use the budget to employ a student for 3-4 months.
-In this time the student will assist in extracting light curves from the FFIs,
-measuring rotation periods and building a rotation period catalog.
+In this time the student will assist in extracting light curves from the
+\tess\ FFIs, measuring rotation periods and building a rotation period
+catalog.
 The student will also be involved in the scientific project of their choosing:

### 2017-10-05 ms.tex
editing on the train

@@
 exoplanet search, we include a list of them with this proposal.
-The stars in these pairs are assumed to be recently disrupted binaries so we
-expect them to have the same age; measuring their rotation periods will
+The stars in these pairs are likely recently disrupted binaries so we
+expect them to have the same age and measuring their rotation periods will
 therefore provide a test of gyrochronology.
@@
 This latter approach could involve either computing an autocorrelation
-function \citep{mcquillan2013}, or fitting a Gaussian process to the time
-series \citep{angus2017, foreman-mackey2017}.
+function \citep[\eg][]{mcquillan2013}, or fitting a Gaussian process to the
+time series \citep{angus2017, foreman-mackey2017}.
 In cases where signals have long periods and low amplitudes, care is needed to
@@
 This will be achieved by fitting for the systematic effects in the light curve
-using the \textsf{everest} model simultaneously with a Gaussian Process model
+using the \textsf{everest} model simultaneously with a Gaussian process model
 for the astrophysical variation.
@@
 preserve stellar variability signals and we will use the \textsf{celerite}
-algorithm \citep{dfm2017} to scale the computations to the size of TESS FFI
-datasets.
+algorithm \citep{foreman-mackey2017} to scale the computations to the size of
+TESS FFI datasets.
 The precision of existing photometric de-trending methods degrades in crowded
@@
 astrophysical variability in each pixel.
-We will tune this method preserve rotation signals and apply it to detect
+We will tune this method to preserve rotation signals and apply it to detect
 rotation across the FFIs.
@@
 calculating a simple (and computationally fast) Lomb-Scargle periodogram for
-the low priority objects and running a full probabilistic period analysis on
-the high priority targets.
+the low priority objects and running a full probabilistic, Gaussian process
+period analysis on the high priority targets.
 High priority targets include: comoving pairs, stars in the CVZs (especially
@@
 techniques.
-Finally, we will release a catalog of rotation periods for two-minute cadence
-targets.
+Finally, we will release a catalog of rotation periods for all two-minute
+cadence targets.
 The full catalog will likely contain several million rotation periods.
+We will release code for inferring stellar ages from rotation periods, based
+on a new gyrochronology relation calibrated using \TESS\HERMES spectra, \Gaia\
+kinematics and comoving pairs.
 We will also release an open source python package for measuring rotation
@@
 \paragraph{Budget Justification}
-PI Angus intends to use the budget to employ a student for 3-4 months.
+PI Angus intends to use the budget to employ an undergraduate student for 3-4
+months.
 In this time the student will assist in extracting light curves from the

### 2017-10-05 ms.tex
more small tweaks

@@
 
-% Still to do: estimate the number of 16th mag stars and brighter in the whole
-% sky.
-% Estimate computation power available.
-% Come up with a better target and priorities list.
+% Still to do: estimate the number of 16th mag stars and brighter in the TESS
+% southern sky.
+% Figure out how to incorporate the target list.
 
@@
 Of the measurable properties for a large ensemble of field stars, rotation
-periods contain the most information about stellar age, and provide the best
+periods contain the most information about stellar age and provide the best
 leverage for advancing our knowledge of galactic archeology as well as
@@
 inefficient magnetic braking at old ages in Solar-mass stars.
-{\bf We propose to answer these questions by providing a catalog of
-rotation periods for \TESS\ Full Frame Image (FFI) stars.}
-% We also intend to use gyrochronology in the cases where it {\it is} applicable
-% to learn about the evolution of exoplanets across the galaxy.
+To answer these questions {\bf we propose to provide a catalog of rotation
+periods for \TESS\ Full Frame Image (FFI) stars.}
 
@@
 distribution as a function of galactocentric position.
-For the all-sky analysis we will focus on F and G stars since the 27.5 day
-baseline of \TESS\ fields will limit measureable rotation periods to less than
-around 15 days.
-% With the April 2018 data release from Gaia (DR2), we estimate that we will be
-% able to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$3
-% kpc.
-% To characterize the rotation period bimodality as a function of galactic
-% position, we propose to analyse the light curves of 10,000 FFI and two-minute
-% targets that overlap with the Tycho-Gaia Astrometric Solution (TGAS) catalog,
-% selecting FGKM dwarfs with a range of galactic positions.
-% The FFI targets will be essential in order to reach $\sim$3kpc distances.
-% Of these, the stars lying in the CVZs will be our highest priority.
+For the all-southern-hemisphere  analysis we will focus on F and G stars since
+the 27.4 day baseline of \TESS\ fields will limit measureable rotation periods
+to less than around 15 days.
 The longer time coverage in the CVZz will allow us to study the rotation
@@
 rotation periods for these lower mass stars.
-{\bf Unlike \kepler\ and \ktwo\ light curves, \TESS\ provides the all-sky
+Unlike \kepler\ and \ktwo\ light curves, {\bf \TESS\ provides the all-sky
 coverage necessary to map the period bimodality across the galaxy.}
@@
 for stars that fall in the Continuous Viewing Zones (CVZs).
-Many of these will have spectra from the \TESS --\HERMES\ survey, from which it
-will be possible to calculate isochronal ages, enabling further calibration of
-the gyrochronology relations.
+Many of these will have spectra from the \TESS --\HERMES\ survey, from which
+it will be possible to calculate isochronal ages, enabling further calibration
+of the gyrochronology relations.
+{\bf We propose to calibrate gyrochronology at old ages for FGKM stars using
+the isochronal ages of \TESS--\hermes\ stars.}
 We will also use the rotation periods of TESS field dwarfs over the whole sky
@@
 stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
+{\bf We propose to calibrate the gyrochronology relations at old ages for FGKM
+stars using vertical actions of \Gaia\ DR2 targets in the CVZ.}
 
-In addition, we will use the rotation periods of comoving stars identified in
-the first \Gaia\ data release \citep{oh2016} to quantify the accuracy and
-precision of gyrochronology.
-We have identified 38 stars in 19 pairs with \Gaia\ proper motions and radial
-velocities from \RAVE\ that are observable by \TESS, using {\tt tvguide}.
-Of these, 34 stars in 17 pairs are in the CTL list.
+We have identified 34 stars in 17 pairs with \Gaia\ proper motions and radial
+velocities from \RAVE\ that are observable by \TESS
+% \footnote{Based on {\tt tvguide}.}
+and are already in the Candidate Target List.
 Since these stars would make interesting targets for both asteroseismology and
@@
 The stars in these pairs are likely recently disrupted binaries so we
-expect them to have the same age and measuring their rotation periods will
+expect them to have similar ages and measuring their rotation periods will
 therefore provide a test of gyrochronology.
+{\bf We will use the rotation periods of comoving stars identified in the
+first \Gaia\ data release \citep{oh2016} to quantify the accuracy and
+precision of gyrochronology.}
 
@@
 The student will also be involved in the scientific project of their choosing:
-either the rotation period bimodality, gyrochronology of low-mass stars, or
+either the rotation period bimodality, gyrochronology of FGKM stars, or
 gyrochronology of comoving pairs.

### 2017-10-05 ms.tex
more small changes

@@
 \paragraph{Expected Impact}
-Initially, we will provide rotation period posterior probability density
-functions and best fit estimates for around 30,000 high priority FFI targets.
-We will then produce a catalog of rotation periods for lower priority FFI
-stars using a combination of Lomb-Scargle periodogram and autocorrelation
-techniques.
+We will provide rotation period posterior probability density functions and
+best fit estimates for around 30,000 high priority FFI targets and a catalog
+of rotation periods for lower priority FFI stars using a combination of
+Lomb-Scargle periodogram and autocorrelation techniques.
 Finally, we will release a catalog of rotation periods for all two-minute

### 2017-10-06 ms.tex
adding Derek's corrections

@@
 \documentclass[letterpaper,12pt,preprint]{hack_aastex}
-
-% Still to do: estimate the number of 16th mag stars and brighter in the TESS
-% southern sky.
-% Figure out how to incorporate the target list.
-
-\input{dfm_stylez}
-\pagestyle{myheadings}
-\markright{\textsf{\footnotesize %
-    Investigating rotational evolution with TESS FFI field dwarfs}}
-
-\bibliographystyle{aasjournal}
-
 \newcommand{\Kepler}{{\it Kepler}}
@@
 \newcommand{\numax}{$\nu_{\mathrm{max}}$}
-
 \newcommand{\racomment}[1]{{\color{red}#1}}
 
-\newcommand{\columbia}{1}
-\newcommand{\ww}{2}
-\newcommand{\cca}{3}
-\newcommand{\florida}{4}
-\newcommand{\princeton}{5}
-\newcommand{\nsf}{4}
-\newcommand{\simons}{2}
-\newcommand{\hubble}{7}
+% Still to do: estimate the number of 16th mag stars and brighter in the TESS
+% southern sky. ~ 15 million, call it 10.
+% Figure out how to incorporate the target list.
+% Get detection probabilities for stars on the list
+% Apply Derek's corrections
+% Calculate the distance at which you get a ~ 15th mag at ~1kpc
+% cite the sullivan paper.
+% Mention vetting.
+
+\input{dfm_stylez}
+\pagestyle{myheadings}
+\markright{\textsf{\footnotesize %
+    Investigating rotational evolution with \TESS\ FFI field dwarfs}}
+
+\bibliographystyle{aasjournal}
 
@@
 
-\title{Investigating magnetic dynamo evolution with TESS field dwarfs}
+\title{Investigating magnetic dynamo evolution with \TESS\ field dwarfs}
 
@@
 Cool spots on the star's surface rotate in-to and out-of view, creating small
-amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
+amplitude ($\pm$ 1\% or less) quasi-periodic changes in the stellar
+brightness.
 Based on results from \Kepler, using precise light curves available from the
@@
 Secondly, a transitioning magnetic dynamo appears to be responsible for
-inefficient magnetic braking at old ages in Solar-mass stars.
+inefficient magnetic braking at old ages in Solar-mass stars
+\citep{vansaders2016}.
 To answer these questions {\bf we propose to provide a catalog of rotation
@@
 Recently, Co-I Davenport found this period bimodality extends throughout all
-masses in the Kepler rotation sample for nearby stars \citep{davenport2017}.
+masses in the \Kepler\ rotation sample for nearby stars \citep{davenport2017}.
 This feature could either reflect a drop in the star formation rate around 600
@@
 spin-down evolution for low-mass stars.
-The TESS FFI targets will provide the ideal dataset to test these two
+The \TESS\ FFI targets will provide the ideal dataset to test these two
 scenarios explaining the appearance of a period bimodality.
@@
 with galactic position.
-We will match the TESS FFI targets to the upcoming second data
+We will match the \TESS\ FFI targets to the upcoming second data
 release from the Gaia mission \citep{perryman2001} to map the rotation period
@@
 For the all-southern-hemisphere  analysis we will focus on F and G stars since
-the 27.4 day baseline of \TESS\ fields will limit measureable rotation periods
+the 27.4 day baseline of \TESS\ fields will limit measurable rotation periods
 to less than around 15 days.
-The longer time coverage in the CVZz will allow us to study the rotation
+The longer time coverage in the CVZ will allow us to study the rotation
 period bimodality in K and M dwarfs, since the bimodality appears at longer
@@
     Rotation period as a function of effective
-temperature for the full McQuillan et al., (2014) Kepler sample in black, and
+temperature for the full McQuillan et al., (2014) \Kepler\ sample in black, and
 the subset of these stars that also feature in the TGAS Gaia DR1 catalogue in
@@
 Although the classical spin-down law of \citet{skumanich1972} (Period
-$\propto$ Age$^1/2$) holds for all open clusters with measured periods, it does
+$\propto$ Age$^{1/2}$) holds for all open clusters with measured periods, it does
 not appear to describe old field stars \citep{angus2015, van-saders2016}.
@@
 the isochronal ages of \TESS--\hermes\ stars.}
-We will also use the rotation periods of TESS field dwarfs over the whole sky
+We will also use the rotation periods of \TESS\ field dwarfs over the whole sky
 to test the gyrochronology relations using galactic kinematics.
@@
 These parameters provide the information necessary to calculate galactocentric
-positions and action angles of the stars, both of which are age indicators.
+positions and action-angle coordinates of the stars, both of which are age
+indicators.
 Since one of the few ways to accurately age-date fully convective, late M
@@
 
-We have identified 34 stars in 17 pairs with \Gaia\ proper motions and radial
-velocities from \RAVE\ that are observable by \TESS
-% \footnote{Based on {\tt tvguide}.}
-and are already in the Candidate Target List.
+\citet{oh2016} identified over 10,000 comoving pairs of stars using proper
+motions from the first \Gaia\ data release.
+Of these, 34 dwarfs and subgiants in 17 pairs have \RAVE\ radial velocities
+consistent with being comoving and are already in the Candidate Target List.
 Since these stars would make interesting targets for both asteroseismology and
-exoplanet search, we include a list of them with this proposal.
+exoplanet search, we include a list with this proposal.
 The stars in these pairs are likely recently disrupted binaries so we
@@
 therefore provide a test of gyrochronology.
-{\bf We will use the rotation periods of comoving stars identified in the
-first \Gaia\ data release \citep{oh2016} to quantify the accuracy and
-precision of gyrochronology.}
+An additional 5000 bright comoving pair candidates lie in the southern
+ecliptic hemisphere and will be ideal FFI targets for gyrochronology studies.
+{\bf We will use the rotation periods of comoving stars to quantify the
+accuracy and precision of gyrochronology.}
 
@@
 We will apply two complementary methods to extract and calibrate light curves
-from the TESS FFIs.
+from the \TESS\ FFIs.
 First, for bright or isolated targets, we will follow \citet{montet2017} to
@@
 algorithm \citep{foreman-mackey2017} to scale the computations to the size of
-TESS FFI datasets.
+\TESS\ FFI datasets.
 The precision of existing photometric de-trending methods degrades in crowded
@@
 We will provide rotation period posterior probability density functions and
-best fit estimates for around 30,000 high priority FFI targets and a catalog
-of rotation periods for lower priority FFI stars using a combination of
+best fit estimates for around 30,000 high priority FFI targets rotation
+periods point estimates for lower priority FFI stars using a combination of
 Lomb-Scargle periodogram and autocorrelation techniques.
-Finally, we will release a catalog of rotation periods for all two-minute
-cadence targets.
-The full catalog will likely contain several million rotation periods.
+% Finally, we will release a catalog of rotation periods for all two-minute
+% cadence targets.
+The full catalog will contain several million rotation periods.
 We will release code for inferring stellar ages from rotation periods, based
-on a new gyrochronology relation calibrated using \TESS\HERMES spectra, \Gaia\
-kinematics and comoving pairs.
+on a new gyrochronology relation calibrated using \TESS -- \HERMES spectra,
+\Gaia\ kinematics and comoving pairs.
 We will also release an open source python package for measuring rotation
-periods of both FFI and two-minute cadence targets.
+periods of \tess\ FFI targets.
 

### 2017-10-06 ms.tex
talk about vetting

@@
 % southern sky. ~ 15 million, call it 10.
+% Cite the sullivan paper.
 % Figure out how to incorporate the target list.
 % Get detection probabilities for stars on the list
-% Apply Derek's corrections
 % Calculate the distance at which you get a ~ 15th mag at ~1kpc
-% cite the sullivan paper.
 % Mention vetting.
@@
 to less than around 15 days.
-The longer time coverage in the CVZ will allow us to study the rotation
-period bimodality in K and M dwarfs, since the bimodality appears at longer
-rotation periods for these lower mass stars.
+The longer time coverage in the Continuous Viewing Zone (CVZ) will allow us to
+study the rotation period bimodality in K and M dwarfs, since the bimodality
+appears at longer rotation periods for these lower mass stars.
 Unlike \kepler\ and \ktwo\ light curves, {\bf \TESS\ provides the all-sky
@@
 Of particular interest are the potentially long rotation periods measurable
-for stars that fall in the Continuous Viewing Zones (CVZs).
+for stars that fall in the CVZ.
 Many of these will have spectra from the \TESS --\HERMES\ survey, from which
@@
 
-The full set of available \TESS\ FFI light curves will be exceedingly large
-and we intend to prioritize certain groups of stars for period analysis,
+Based on \citet{sullivan2015}, we expect around 10 million \TESS\ FFI
+targets brighter than 15th magnitude in the southern ecliptic hemisphere.
+We intend to prioritize certain groups of stars for period analysis,
 calculating a simple (and computationally fast) Lomb-Scargle periodogram for
@@
 period analysis on the high priority targets.
-High priority targets include: comoving pairs, stars in the CVZs (especially
-those with spectra), stars in TGAS, stars at large distances, plus any
-potential asteroseismic dwarfs and exoplanet hosts.
+High priority targets will include: comoving pairs, stars in the CVZ
+(especially those with spectra), stars in TGAS, stars at large distances, plus
+any potential asteroseismic dwarfs and exoplanet hosts.
 Based on the expected photometric precision of \tess\ \citep{sullivan2015}, we
-expect to be able to measure rotation periods for FFI stars down to 16th
+expect to be able to measure rotation periods for FFI stars down to 15th
 magnitude.
@@
 inactive.
+We will vet incorrect rotation period measurements using a combination of a
+Lomb-Scargle periodogram and an autocorrelation function by
+training on a set of light curves with a range of variability amplitudes and
+periods.
+However, based on the high success rates of a Lomb-Scargle periodogram method
+with high-pass filtered light curve and the Gaussian process method
+demonstrated in \citet{angus2017}, we expect a low rate of incorrect rotation
+periods.
 

### 2017-10-06 ms.tex
estimating number of rotation periods

@@
 
-% Still to do: estimate the number of 16th mag stars and brighter in the TESS
-% southern sky. ~ 15 million, call it 10.
-% Cite the sullivan paper.
 % Figure out how to incorporate the target list.
@@
 % Calculate the distance at which you get a ~ 15th mag at ~1kpc
-% Mention vetting.
+% Shorten.
 
@@
 inefficient magnetic braking at old ages in Solar-mass stars
-\citep{vansaders2016}.
+\citep{van-saders2016}.
 To answer these questions {\bf we propose to provide a catalog of rotation
@@
 to test the gyrochronology relations using galactic kinematics.
-Many stars observable by \tess\ will have proper motions, parallaxes,
-positions and radial velocities published in the second \Gaia\ data release.
+Many \tess\ FFI targets will have proper motions, parallaxes, positions and
+radial velocities published in the second \Gaia\ data release.
 These parameters provide the information necessary to calculate galactocentric
@@
 any potential asteroseismic dwarfs and exoplanet hosts.
+The number of high priority targets with probabilistic rotation periods will
+be limited by computation power, but we conservatively estimate that we can
+produce full rotation period posterior PDFs for $\sim$ 30,000 stars.
 Based on the expected photometric precision of \tess\ \citep{sullivan2015}, we
@@
 We note, however that not all stars will have measurable rotation periods,
-even if they are bright enough --- not all stars show rotational variability
-in their light curves either because they are pole-on or because they are
-inactive.
+even if they are bright enough not all stars show rotational variability in
+their light curves either because they are pole-on or inactive.
 We will vet incorrect rotation period measurements using a combination of a
@@
 However, based on the high success rates of a Lomb-Scargle periodogram method
-with high-pass filtered light curve and the Gaussian process method
+with high-pass filtered light curve and the Gaussian process method,
 demonstrated in \citet{angus2017}, we expect a low rate of incorrect rotation
@@
 We will provide rotation period posterior probability density functions and
-best fit estimates for around 30,000 high priority FFI targets rotation
-periods point estimates for lower priority FFI stars using a combination of
-Lomb-Scargle periodogram and autocorrelation techniques.
-% Finally, we will release a catalog of rotation periods for all two-minute
-% cadence targets.
-The full catalog will contain several million rotation periods.
-We will release code for inferring stellar ages from rotation periods, based
-on a new gyrochronology relation calibrated using \TESS -- \HERMES spectra,
-\Gaia\ kinematics and comoving pairs.
-We will also release an open source python package for measuring rotation
-periods of \tess\ FFI targets.
+best fit estimates for $\sim$ 30,000 high priority FFI targets rotation
+periods and point estimates for $\sim$ 10 million lower priority FFI stars
+using a combination of Lomb-Scargle periodogram and autocorrelation
+techniques.
+We will release an open source python package for measuring rotation periods
+of \tess\ FFI targets plus code for inferring stellar ages from rotation
+periods, based on a new gyrochronology relation calibrated using \TESS --
+\HERMES spectra, \Gaia\ kinematics and comoving pairs.
 
 \paragraph{Budget Justification}
-PI Angus intends to use the budget to employ an undergraduate student for 3-4
-months.
-In this time the student will assist in extracting light curves from the
-\tess\ FFIs, measuring rotation periods and building a rotation period
-catalog.
-The student will also be involved in the scientific project of their choosing:
-either the rotation period bimodality, gyrochronology of FGKM stars, or
-gyrochronology of comoving pairs.
-The remaining budget will be used for traveling to meet with other Co-Is in
-order to perform this analysis.
+PI Angus will employ two undergraduate students for 3-4 months who will assist
+in extracting light curves from the \tess\ FFIs, measuring rotation periods
+and building a rotation period catalog.
+These students will also be involved in the scientific project of their
+choosing: either the rotation period bimodality, gyrochronology of FGKM stars,
+or gyrochronology of comoving pairs.
+The remaining budget will be used for a bringing all Co-Is together in order
+to make and execute analysis plans and for presenting the results at
+conferences.
 \paragraph{References}

### 2017-10-06 ms.tex
addding references

@@
 \citep{van-saders2016}.
-To answer these questions {\bf we propose to provide a catalog of rotation
-periods for \TESS\ Full Frame Image (FFI) stars.}
+To answer these questions {\bf we propose to provide a catalog of $\sim$ 1
+million rotation periods for \TESS\ Full Frame Image (FFI) stars.}
 
@@
 motions from the first \Gaia\ data release.
-Of these, 34 dwarfs and subgiants in 17 pairs have \RAVE\ radial velocities
+Of these, 18 dwarfs and subgiants in 9 pairs have \RAVE\ radial velocities
 consistent with being comoving and are already in the Candidate Target List.
+Following \citet{campante2017}, these stars have oscillation detection
+probabilities greater than 50\%, based on their \RAVE-on T$_{\mathrm{eff}}$,
+$\log g$s and radii \citep{casey2017}.
 Since these stars would make interesting targets for both asteroseismology and
@@
 periods.
-However, based on the high success rates of a Lomb-Scargle periodogram method
-with high-pass filtered light curve and the Gaussian process method,
-demonstrated in \citet{angus2017}, we expect a low rate of incorrect rotation
-periods.
+Based on the fraction of stars with measurable rotation periods in
+\citet{mcquillan2014}, and the more restrictive baseline of \tess\ relative to
+\kepler, we predict that the final catalog will contain around one million
+rotation periods.
 

### 2017-10-06 ms.tex
editing first section

@@
 brightness.
-Based on results from \Kepler, using precise light curves available from the
-\TESS\ mission, we expect to measure the rotation periods of around 20\% of FG
-stars and 80\% of KM stars in the \tess\ Full Frame Images (FFIs)\footnote{The
-difference between these percentages comes from a dramatic increase in
-magnetic activity, and therefore more easily measured rotation periods,
-towards lower masses.}.
-For many of these stars it will be possible to infer an age from their
-rotation periods via gyrochronology.
+Thousands of photometric rotation periods have been measured from \Kepler\ and
+\ktwo\ light curves, but these surveys focused on small areas of the sky.
+\tess\ will open unexplored parts of the galaxy to rotational evolution
+studies.
 Some outstanding mysteries regarding the nature of magnetic braking were

### 2017-10-06 ms.tex
some more editing and shortening

@@
 amplitude ($\pm$ 1\% or less) quasi-periodic changes in the stellar
-brightness.
+brightness from which a rotation period can be measured.
 Thousands of photometric rotation periods have been measured from \Kepler\ and
-\ktwo\ light curves, but these surveys focused on small areas of the sky.
-\tess\ will open unexplored parts of the galaxy to rotational evolution
-studies.
+\ktwo\ light curves this way, but these surveys focused on small areas of the
+sky whereas \tess\ will open unexplored parts of the galaxy to rotational
+evolution studies.
 Some outstanding mysteries regarding the nature of magnetic braking were
@@
 \citep{van-saders2016}.
-To answer these questions {\bf we propose to provide a catalog of $\sim$ 1
-million rotation periods for \TESS\ Full Frame Image (FFI) stars.}
+To shed light on this behavior {\bf we propose to provide a catalog of $\sim$
+1 million rotation periods for \TESS\ Full Frame Image (FFI) stars, to map the
+rotation period bimodality as a function of galactic position and to calibrate
+gyrochronology using \tess--\hermes\ spectra, \gaia\ kinematics and \gaia\
+comoving pairs.}
 
@@
 The \TESS\ FFI targets will provide the ideal dataset to test these two
-scenarios explaining the appearance of a period bimodality.
-If the bimodal period distribution reflects a discontinuous age distribution,
-the feature should be local and may disappear at greater distances or along
+scenarios.
+If the period bimodality reflects a discontinuous age distribution, the
+feature should be local and may disappear at greater distances or along
 different lines of sight.
@@
 release from the Gaia mission \citep{perryman2001} to map the rotation period
-distribution as a function of galactocentric position.
-For the all-southern-hemisphere  analysis we will focus on F and G stars since
+distribution as a function of galactic position.
+Outside the Continuous Viewing Zone (CVZ) we will focus on F and G stars since
 the 27.4 day baseline of \TESS\ fields will limit measurable rotation periods
 to less than around 15 days.
-The longer time coverage in the Continuous Viewing Zone (CVZ) will allow us to
-study the rotation period bimodality in K and M dwarfs, since the bimodality
-appears at longer rotation periods for these lower mass stars.
+The longer time coverage in the CVZ will allow us to study the rotation period
+bimodality in K and M dwarfs too, since the bimodality appears at longer
+rotation periods for these lower mass stars.
 Unlike \kepler\ and \ktwo\ light curves, {\bf \TESS\ provides the all-sky
@@
 Although the classical spin-down law of \citet{skumanich1972} (Period
-$\propto$ Age$^{1/2}$) holds for all open clusters with measured periods, it does
-not appear to describe old field stars \citep{angus2015, van-saders2016}.
+$\propto$ Age$^{1/2}$) holds for all open clusters with measured periods, it
+does not appear to describe old field stars \citep{angus2015, van-saders2016}.
 \citet{van-saders2016} found that including a transition to a weakened magnetic
@@
 than the Sun relatively unexplored.
-\TESS\ will provide thousands of new rotation periods, with the maximum
-measurable period strongly depending on the length of time it is observed for.
-Of particular interest are the potentially long rotation periods measurable
-for stars that fall in the CVZ.
+\TESS\ will provide thousands of new {\it long } rotation periods for stars
+that fall in the CVZ.
 Many of these will have spectra from the \TESS --\HERMES\ survey, from which
@@
 indicators.
+{\bf We propose to calibrate the gyrochronology relations at old ages for FGKM
+stars using vertical actions of \Gaia\ DR2 targets.}
 Since one of the few ways to accurately age-date fully convective, late M
@@
 stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
-{\bf We propose to calibrate the gyrochronology relations at old ages for FGKM
-stars using vertical actions of \Gaia\ DR2 targets in the CVZ.}
 
@@
 consistent with being comoving and are already in the Candidate Target List.
-Following \citet{campante2017}, these stars have oscillation detection
-probabilities greater than 50\%, based on their \RAVE-on T$_{\mathrm{eff}}$,
-$\log g$s and radii \citep{casey2017}.
+Following \citet{campante2016} we calculate that these stars have oscillation
+detection probabilities greater than 50\%, based on their \RAVE-on
+T$_{\mathrm{eff}}$, $\log g$ and radii \citep{casey2017}.
 Since these stars would make interesting targets for both asteroseismology and
@@
 (especially those with spectra), stars in TGAS, stars at large distances, plus
-any potential asteroseismic dwarfs and exoplanet hosts.
+potential asteroseismic dwarfs and exoplanet hosts.
 The number of high priority targets with probabilistic rotation periods will
@@
 magnitude.
-We note, however that not all stars will have measurable rotation periods,
+We note however that not all stars will have measurable rotation periods:
 even if they are bright enough not all stars show rotational variability in
@@
 training on a set of light curves with a range of variability amplitudes and
-periods.
+known periods.
 Based on the fraction of stars with measurable rotation periods in
@@
 
-\ssfigure{Kalesalad.png}{0.5}{%
+\ssfigure{Kalesalad.png}{0.3}{%
 All stars observed during K2 campaign 4, plotted according to their
@@
 best fit estimates for $\sim$ 30,000 high priority FFI targets rotation
-periods and point estimates for $\sim$ 10 million lower priority FFI stars
+periods and point estimates for $\sim$ 1 million lower priority FFI stars
 using a combination of Lomb-Scargle periodogram and autocorrelation
@@
 of \tess\ FFI targets plus code for inferring stellar ages from rotation
-periods, based on a new gyrochronology relation calibrated using \TESS --
-\HERMES spectra, \Gaia\ kinematics and comoving pairs.
+periods, based on a new gyrochronology relation calibrated using \TESS--
+\HERMES\ spectra, \Gaia\ kinematics and comoving pairs.
 
@@
 PI Angus will employ two undergraduate students for 3-4 months who will assist
-in extracting light curves from the \tess\ FFIs, measuring rotation periods
-and building a rotation period catalog.
+in extracting light curves from the \tess\ FFIs and building a rotation period
+catalog.
 These students will also be involved in the scientific project of their
@@
 or gyrochronology of comoving pairs.
-The remaining budget will be used for a bringing all Co-Is together in order
+The remaining budget will be used for bringing all Co-Is together in order
 to make and execute analysis plans and for presenting the results at

### 2017-10-06 ms.tex
adding real number of stars

@@
 motions from the first \Gaia\ data release.
-Of these, 18 dwarfs and subgiants in 9 pairs have \RAVE\ radial velocities
+Of these, 8 dwarfs and subgiants in 4 pairs have \RAVE\ radial velocities
 consistent with being comoving and are already in the Candidate Target List.
 Following \citet{campante2016} we calculate that these stars have oscillation
-detection probabilities greater than 50\%, based on their \RAVE-on
-T$_{\mathrm{eff}}$, $\log g$ and radii \citep{casey2017}.
+detection probabilities greater than 0.85, based on their \RAVE-on
+T$_{\mathrm{eff}}$, $\log g$ and radii \citep{casey2017} and their observation
+baseline, from {\tt tvguide}.
 Since these stars would make interesting targets for both asteroseismology and

### 2017-10-06 ms.tex
edits

@@
 detection probabilities greater than 0.85, based on their \RAVE-on
-T$_{\mathrm{eff}}$, $\log g$ and radii \citep{casey2017} and their observation
-baseline, from {\tt tvguide}.
+T$_{\mathrm{eff}}$, $\log g$ and radii \citep{casey2017} and their {\tt
+tvguide} \citep{tvguide} observation baseline.
 Since these stars would make interesting targets for both asteroseismology and
 exoplanet search, we include a list with this proposal.
-The stars in these pairs are likely recently disrupted binaries so we
-expect them to have similar ages and measuring their rotation periods will
-therefore provide a test of gyrochronology.
+The stars in these pairs are likely recently disrupted binaries so we expect
+them to have similar ages and measuring their rotation periods
+will therefore provide a test of gyrochronology, even in cases where
+asteroseismic ages are unavailable.
 An additional 5000 bright comoving pair candidates lie in the southern

### 2017-10-06 ms.tex
blah

@@
 From Davenport (2017).
-    Rotation period as a function of effective
-temperature for the full McQuillan et al., (2014) \Kepler\ sample in black, and
-the subset of these stars that also feature in the TGAS Gaia DR1 catalogue in
-blue. Contaminating giants have been removed from the blue sample and the
+    Rotation period as a function of effective temperature for the full
+McQuillan et al., (2014) \Kepler\ sample in black, and the subset of these
+stars that also feature in the Tycho-Gaia Astrometric Solution (TGAS) Gaia DR1
+catalog in blue. Contaminating giants have been removed from the blue sample
+and the
 rotation period bimodality is revealed to exist across all temperatures shown.
@@
 \TESS\ will provide thousands of new {\it long } rotation periods for stars
-that fall in the CVZ.
+that fall in the CVZs.
 Many of these will have spectra from the \TESS --\HERMES\ survey, from which
@@
 their light curves either because they are pole-on or inactive.
-We will vet incorrect rotation period measurements using a combination of a
-Lomb-Scargle periodogram and an autocorrelation function by
-training on a set of light curves with a range of variability amplitudes and
-known periods.
+We will vet incorrect rotation period measurements by training our algorithms
+on a set of light curves with a range of variability amplitudes and known
+periods.
 Based on the fraction of stars with measurable rotation periods in
